---
title: "R Practice Answers"
author: "Ryan Safner"
date: "September 12, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Set Up 

Before we begin, start a new file with \texttt{File} $\rightarrow$ \texttt{New File} $\rightarrow$ \texttt{R Script}. As you work through this sheet in the console in `R`, also add (copy/paste) your commands that work into this new file. At the end, save it, and run to execute all of your commands at once. 

## Creating Objects

### 1. Create a vector called "me" with two objects, your first name, and your last name. Then call the vector to inspect it. Confirm it is a character class vector.
```{r}
me<-c("Ryan","Safner")
me
class(me)
```

### 2. Create a vector called "x" with all the even integers from 2 to 10. 
```{r}
x<-c(2,4,6,8,10)
```

### 3. Find the mean of x with `mean()`
```{r}
mean(x)
```

### 4. Now take the following pdf of random variable $Y$:

\begin{table}[h!]
\centering
  \begin{tabular}{ll}
  $y_i$ & $p_i$\\ \hline
  2 & 0.50\\
  4 & 0.25\\
  6 & 0.25\\ \hline
  \end{tabular}
\end{table}

Calculate the standard deviation "manually" using our table method. You can look at the source code of Lecture 4 for my example.  

a. Creating two vectors, one called $y.i$ and one called $p.i$, with the data above. 
```{r}
y.i<-c(2,4,6)
p.i<-c(0.5,0.25,0.25)
```

b. Merge them into a data frame called $rv$ with `data.frame(y.i,p.i)`. Call `rv` to inspect it. 
```{r}
rv<-data.frame(y.i,p.i)
rv
```

c. Find the expected value of $Y$ by taking the sum of each value of `y.i` multiplied by `p.i` with the `sum()` command. 
```{r}
sum(rv$y.i*rv$p.i)
```

d. Creating a new column in `rv` called `deviations`, where you subtract the mean from each `y.i` value. Call `rv` again to make sure it's now there. 
```{r}
rv$deviations<-(rv$y.i-3.5)
rv
```

e. Create another column in `rv` called `devsq`, where you square the deviations from part d. Call `rv` again to make sure it's now there.
```{r}
rv$devsq<-(rv$deviations^2)
rv
```

f. Now add another column in `rv` called `weighteddevsq`, where you multiply the squared deviations in part e. by the associated probability `p.i`. Call `rv` again to make sure it's now there.
```{r}
rv$weighteddevsq<-(rv$devsq*rv$p.i)
rv
```

g. Finally, take the sum of `weighteddevsq` to get variance. Square root this to get standard deviation.
```{r}
sum(rv$weighteddevsq)
sqrt(sum(rv$weighteddevsq))
```

### 5. The mean height of adults is 65 inches, with a standard deviation of 4 inches. Use the normal distribution to find the probabilities of the following scenarios: 

a. Find the probability of someone being *at least* 60 inches tall using `pnorm()`.
```{r}
pnorm(60, mean=65, sd=4, lower.tail=FALSE) 
```


b. Find the probability of someone being *at most* 60 inches tall. 
```{r}
pnorm(60, mean=65, sd=4, lower.tail=TRUE) 
```

c. Find the probability of someone being between 61 and 69 inches tall. Why is this number familiar? 
```{r}
pnorm(69, mean=65, sd=4, lower.tail=TRUE)-pnorm(61, mean=65, sd=4, lower.tail=TRUE)
```

d. Find the probability of someone being between 57 and 73 inches tall. Why is this number familiar? 

```{r}
pnorm(73, mean=65, sd=4, lower.tail=TRUE)-pnorm(57, mean=65, sd=4, lower.tail=TRUE)
```

## Playing with a Data Set 

For the following questions, use the \texttt{diamonds} dataset, included as part of `ggplot2`.

### 1. Install `ggplot2`
```
install.packages("ggplot2")
```

### 2. Load `ggplot2` with the `library()` command

```{r}
library(ggplot2)
```

### 3. Get the structure of the diamonds data.frame. What are the different variables and what kind of data does each contain? 
```{r}
str(diamonds)
```

### 4. Get summary statistics for `carat`, `depth`, `table`, and `price`

```{r}
summary(diamonds$carat)
summary(diamonds$depth)
summary(diamonds$table)
summary(diamonds$price)
```

### 5. `color`, `cut`, and `clarity` are categorical variables (factors). Use the `table()` command to generate frequency tables for each.

```{r}
table(diamonds$cut)
table(diamonds$color)
table(diamonds$clarity)
```
Note, you can also use `summary()` to get the counts of each category. 

### 6. Now rerun the `summary()` command on the entire data frame
```{r}
summary(diamonds)
```

### 7. Plot a histogram of price. 
```{r}
hist(diamonds$price)
```

### 8. Plot a boxplot of price by diamond color.
```{r}
boxplot(price~cut,data=diamonds)
```

## Execute your R Script

Save the R Script you created at the beginning and (hopefully) have been pasting all of your valid commands to. This creates a *.R* file wherever you choose to save it to. Now looking at the file in the upper left pane of *R Studio* look for the button in the upper right corner that says **Run**. Sit back and watch `R` redo everything you've carefully worked on, all at once. 