---
title: "Lecture 19: Difference-in-Difference Models"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "December 5, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(tidyverse) 
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, dev.args=list(bg="transparent"))
```

# Difference-in-Difference Models

### Difference-in-Difference Models 

- Often, we want to examine the consequences of a change, such as a law or policy
    - e.g. States that implemented law $X$ saw a change in $Y$
    - **Treatment**: States that implement law $X$
    - **Control**: States that did not implement law $X$
    - If we have **panel data** with observations for all states before *and* after the change:
- Simple logic: compare difference in outcomes of treatment group (before and after treatment) with those of non-treated group (before and after same treatment period)

### Difference-in-Difference Models II

\begin{figure}
  \includegraphics[height=2.5in]{dndhot}
\end{figure}

### Difference-in-Difference Models III

- The \alert{difference-in-difference model} 	(aka \alert{"diff-in-diff"}" or \alert{"DND"}) identifies treatment effect by differencing the difference pre- and post-treatment between treatment and control groups
$$\widehat{Y_{it}}=\beta_0+\beta_1 \text{Treated}_i +\beta_2 \text{After}_{t}+\beta_3 (\text{Treated}_i * \text{After}_{t})+\epsilon_{it}$$
    - $Treated_i= \begin{cases}1 \text{ if } i \text{ is in treatment group}\\ 0 \text{ if } i \text{ is not in treatment group}\end{cases}$
    - $After_t= \begin{cases}1 \text{ if } t \text{ is after treatment period}\\ 0 \text{ if } t \text{ is before treatment period}\end{cases}$

\vspace{-0.5cm}

\onslide<4->\begin{table}\small 
		\centering 
		\begin{tabular}{rrrr}
		 & Control & Treatment & \textcolor{red}{Group Diff. ($\Delta Y_i$)}\\ \toprule 
		 Before & $\beta_0$ & $\beta_0+\beta_1$ & \textcolor{red}{$\beta_1$}\\
		 After & $\beta_0+\beta_2$ & $\beta_0+\beta_1+\beta_2+\beta_3$ & \textcolor{red}{$\beta_1+\beta_3$}\\ \midrule 
		 \textcolor{blue}{Time Diff. ($\Delta Y_t$)} & \textcolor{blue}{$\beta_2$}	 & \textcolor{blue}{$\beta_2+\beta_3$} & \textcolor{purple}{$\beta_3$}\\ \bottomrule 
		 & & & \textcolor{purple}{Diff-in-diff ($\Delta_i \Delta_t Y$)}\\ 
		\end{tabular}
	\end{table}
	
### Visualizing Diff-in-Diff
	
$$\widehat{Y_{it}}=\hat{\beta_0}+\hat{\beta_1} \text{Treated}_i +\hat{\beta_2} \text{After}_{t}+\hat{\beta_3} (\text{Treated}_i * \text{After}_{t})+\hat{\epsilon_{it}}$$
	\vspace{-0.5cm}
	\begin{columns}
		\begin{column}[c]{7cm}
	\begin{figure}
				\begin{tikzpicture}[scale=.45]\scriptsize 
				\draw[->] (-1,0) -- (11,0) coordinate (x axis) node[right]{$time$};
 				\draw[->] (-1,0) -- (-1,11) coordinate (y axis) node[above]{$Y$};	
 				\draw<2->[fill=black](2,2)circle(0.1cm)node[below]{$C_1$};
 				\draw<2->[fill=black](8,4)circle(0.1cm)node[above]{$C_2$};
 				\draw<6->[fill=black](2,5)circle(0.1cm)node[above]{$T_1$};
 				\draw<6->[fill=black](8,10)circle(0.1cm)node[above]{$T_2$};
            	\draw<3->[ultra thick, magenta] (2,2)--(8,4);
            	\draw<5->[ultra thick, dotted, magenta] (2,2)--(8,2)--node[right]{$\hat{\beta_2}$}(8,4);
            	\draw<7->[ultra thick, teal] (2,5)node[left]{$\hat{\beta_0}+\hat{\beta_1}$}--(8,10);
            	\draw<9->[ultra thick, dotted, teal] (2,5)--(8,5);
            	\draw<9->[ultra thick, dotted, magenta](8,5)--node[right]{$\hat{\beta_2}$}(8,7);
            	\draw<9->[ultra thick, dotted, magenta](2,5)--(8,7);
            	\draw<10->[ultra thick, blue](8,7)--node[right]{$\hat{\beta_3}$}(8,10);
            	\draw (2,0.25)--(2,-0.25)node[below]{$t_{Before}$};
            	\draw (8,0.25)--(8,-0.25) node[below]{$t_{After}$};
            	\draw<8->[thick, dotted](2,2)--node[left]{$\hat{\beta_1}$}(2,5);
            	\draw<4->[thick, dotted](2,0)--(2,2)node[left]{\textcolor{magenta}{$\hat{\beta_0}$}};
 			\end{tikzpicture}	
	\end{figure}
		\end{column}
		\begin{column}[c]{8cm}
		\begin{itemize}\small 
			\item<7-> \textcolor{teal}{Treatment ($Treated=1$) group}
			\item<3-> \textcolor{magenta}{Control ($Treated=0$) group}
			\item<4->$\hat{\beta_0}$: value of $Y$ for control before treatment 
			\item<8->$\hat{\beta_1}$: difference between treatment and control (before treatment)
			\item<5->$\hat{\beta_2}$: time difference (for control group)
			\item<10->\textcolor{blue}{$\hat{\beta_3}$}: difference-in-difference: effect of treatment    
		\end{itemize}
		\end{column}
	\end{columns}	

### Visualizing and Comparing Group Means

$$\widehat{Y_{it}}=\hat{\beta_0}+\hat{\beta_1} \text{Treated}_i +\hat{\beta_2} \text{After}_{t}+\hat{\beta_3} (\text{Treated}_i * \text{After}_{t})+\hat{\epsilon_{it}}$$
\vspace{-0.5cm}
\begin{columns}
		\begin{column}[c]{7cm}
	\begin{figure}
				\begin{tikzpicture}[scale=.45]\scriptsize 
				\draw[->] (-1,0) -- (11,0) coordinate (x axis) node[right]{$time$};
 				\draw[->] (-1,0) -- (-1,11) coordinate (y axis) node[above]{$Y$};	
 				\draw[fill=black](2,2)circle(0.1cm)node[below]{$C_1$};
 				\draw[fill=black](8,4)circle(0.1cm)node[above]{$C_2$};
 				\draw[fill=black](2,5)circle(0.1cm)node[above]{$T_1$};
 				\draw[fill=black](8,10)circle(0.1cm)node[above]{$T_2$};
            	\draw[ultra thick, magenta] (2,2)--(8,4);
            	\draw[ultra thick, dotted, magenta] (2,2)--(8,2)--(8,4);
            	\draw[ultra thick, teal] (2,5)--(8,10);
            	\draw[ultra thick, dotted, teal] (2,5)--(8,5);
            	\draw[ultra thick, dotted, magenta](8,5)--(8,7);
            	\draw[ultra thick, dotted, magenta](2,5)--(8,7);
            	\draw[ultra thick, teal, dotted](8,7)--(8,10);
            	\draw (2,0.25)--(2,-0.25)node[below]{$t_{Before}$};
            	\draw (8,0.25)--(8,-0.25) node[below]{$t_{After}$};
            	\draw[thick, dotted](2,0)--(2,2);
            	\draw<2-> (2,2)node[left]{$\hat{\beta_0}$};
            	\draw<3-> (8,4)node[right]{$\hat{\beta_0}+\hat{\beta_2}$};
            	\draw<4-> (2,5)node[left]{$\hat{\beta_0}+\hat{\beta_1}$};
            	\draw<5-> (8,10)node[right]{\tiny $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$};
            	\draw<6->[color=blue, thick] (8,7)--(8,10);
 			\end{tikzpicture}	
	\end{figure}
		\end{column}
		\begin{column}[c]{8cm}
		\begin{itemize}\small 
			\item<2-> $Y$ for Control Group Before: $\hat{\beta_0}$
			\item<3-> $Y$ for Control Group After: $\hat{\beta_0}+\hat{\beta_2}$
			\item<4-> $Y$ for Treatment Group Before: $\hat{\beta_0}+\hat{\beta_1}$
			\item<5-> $Y$ for Treatment Group After: $\hat{\beta_0}+\hat{\beta_1}+\hat{\beta_2}+\hat{\beta_3}$
			\item<6-> Treatment Effect: \textcolor{blue}{$\hat{\beta_3}$}
		\end{itemize}
		\end{column}
\end{columns}	

### Comparing Group Means (Again)

$$\widehat{Y_{it}}=\hat{\beta_0}+\hat{\beta_1} \text{Treated}_i +\hat{\beta_2} \text{After}_{t}+\hat{\beta_3} (\text{Treated}_i * \text{After}_{t})+\hat{\epsilon_{it}}$$

\begin{table}\small 
		\centering 
		\begin{tabular}{rrrr}
		 & Control & Treatment & \textcolor{red}{Group Diff. ($\Delta Y_i$)}\\ \toprule 
		 Before & $\beta_0$ & $\beta_0+\beta_1$ & \textcolor{red}{$\beta_1$}\\
		 After & $\beta_0+\beta_2$ & $\beta_0+\beta_1+\beta_2+\beta_3$ & \textcolor{red}{$\beta_1+\beta_3$}\\ \midrule 
		 \textcolor{blue}{Time Diff. ($\Delta Y_t$)} & \textcolor{blue}{$\beta_2$}	 & \textcolor{blue}{$\beta_2+\beta_3$} & \textcolor{purple}{$\beta_3$}\\ \bottomrule 
		 & & & \textcolor{purple}{Diff-in-diff ($\Delta_i \Delta_t Y$)}\\ 
		\end{tabular}
\end{table}

### Key Assumption About Counterfactual

$$\widehat{Y_{it}}=\hat{\beta_0}+\hat{\beta_1} \text{Treated}_i +\hat{\beta_2} \text{After}_{t}+\hat{\beta_3} (\text{Treated}_i * \text{After}_{t})+\hat{\epsilon_{it}}$$
\vspace{-0.5cm}
\begin{columns}
		\begin{column}[c]{7cm}
	\begin{figure}
				\begin{tikzpicture}[scale=.45]\scriptsize 
				\draw[->] (-1,0) -- (11,0) coordinate (x axis) node[right]{$time$};
 				\draw[->] (-1,0) -- (-1,11) coordinate (y axis) node[above]{$Y$};	
 				\draw[fill=black](2,2)circle(0.1cm)node[below]{$C_1$};
 				\draw[fill=black](8,4)circle(0.1cm)node[above]{$C_2$};
 				\draw[fill=black](2,5)circle(0.1cm)node[above]{$T_1$};
 				\draw[fill=black](8,10)circle(0.1cm)node[above]{$T_2$};
            	\draw[ultra thick, magenta] (2,2)--(8,4);
            	\draw[ultra thick, dotted, magenta] (2,2)--(8,2)--(8,4);
            	\draw[ultra thick, teal] (2,5)--(8,10);
            	\draw[ultra thick, dotted, teal] (2,5)--(8,5);
            	\draw[ultra thick, dotted, magenta](8,5)--(8,7);
            	\draw[ultra thick, dotted, magenta](2,5)--(8,7);
            	\draw[ultra thick, teal, dotted](8,7)--(8,10);
            	\draw (2,0.25)--(2,-0.25)node[below]{$t_{Before}$};
            	\draw (8,0.25)--(8,-0.25) node[below]{$t_{After}$};
            	\draw[thick, dotted](2,0)--(2,2);
 			\end{tikzpicture}	
	\end{figure}
		\end{column}
		\begin{column}[c]{8cm}
		\begin{itemize}\small 
			\item<2-> Key assumption in DND models is that the time trend is parallel
			\item<3-> Treatment \& control groups must be similar over time \emph{except for treatment}
			\item<4-> \textcolor{purple}{Counterfactual} if the treatment group were \emph{not} treated, they would change the same as control group over time $\hat{\beta_2}$
		\end{itemize}
		\end{column}
	\end{columns}	

### Key Assumption about Counterfactual II

$$\widehat{Y_{it}}=\hat{\beta_0}+\hat{\beta_1} \text{Treated}_i +\hat{\beta_2} \text{After}_{t}+\hat{\beta_3} (\text{Treated}_i * \text{After}_{t})+\hat{\epsilon_{it}}$$
\vspace{-0.5cm}
\begin{columns}
		\begin{column}[c]{7cm}
	\begin{figure}
				\begin{tikzpicture}[scale=.45]\scriptsize 
				\draw[->] (-1,0) -- (11,0) coordinate (x axis) node[right]{$time$};
 				\draw[->] (-1,0) -- (-1,11) coordinate (y axis) node[above]{$Y$};	
 				\draw[fill=black](2,2)circle(0.1cm)node[below]{$C_1$};
 				\draw[fill=black](8,4)circle(0.1cm)node[above]{$C_2$};
 				\draw[fill=black](2,5)circle(0.1cm)node[above]{$T_1$};
 				\draw[fill=black](8,10)circle(0.1cm)node[above]{$T_2$};
            	\draw[ultra thick, magenta] (2,2)--(8,4);
            	\draw[ultra thick, dotted, magenta] (2,2)--(8,2)--(8,4);
            	\draw[ultra thick, teal] (2,5)--(8,10);
            	\draw[ultra thick, dotted, teal] (2,5)--(8,5);
            	\draw[ultra thick, dotted, magenta](8,5)--(8,5.5);
            	\draw[ultra thick, dotted, magenta](2,5)--(8,5.5);
            	\draw[ultra thick, teal, dotted](8,5.5)--(8,10);
            	\draw (2,0.25)--(2,-0.25)node[below]{$t_{Before}$};
            	\draw (8,0.25)--(8,-0.25) node[below]{$t_{After}$};
            	\draw[thick, dotted](2,0)--(2,2);
 			\end{tikzpicture}	
	\end{figure}
		\end{column}
		\begin{column}[c]{8cm}
		\begin{itemize}\small 
			\item<2-> If time trend is different between treatment and control groups
			\item<3-> Treatment effect may be over/under-estimated!  
		\end{itemize}
		\end{column}
	\end{columns}	

### Diff-in-Diff Example

\begin{example}
	In 1993 Georgia initiated a HOPE scholarship program to let state residents with at least a B average in high school attend public college in Georgia for free. Did it increase college enrollment?  
	\begin{itemize}
		\item<2-> Dynarski, Susan (2000), "Hope for Whom? Financial Aid for the Middle Class and Its Impact on College Attendance" micro-level data on 4,291 young individuals: 
		\begin{itemize}
			\item<3-> InCollege$_{it}=\begin{cases}1 \text{ if } i \text{ is in college during year }t\\ 0 \text{ if } i \text{ is not in college during year }t\\ \end{cases}$
			\item<4-> Georgia$_i=\begin{cases}1 \text{ if } i \text{ is a Georgia resident}\\ 0 \text{ if } i \text{ is not a Georgia resident}\\ \end{cases}$
			\item<5-> After$_t=\begin{cases}1 \text{ if } t \text{ is after 1992}\\ 0 \text{ if } t \text{ is after 1992}\\ \end{cases}$
			\item<6-> \textcolor{magenta}{Note: With a dummy \emph{dependent} variable ($Y$), coefficients estimate the probability $Y=1$, i.e. the \emph{probability} a person is enrolled in college}
		\end{itemize} 
  \end{itemize}
\end{example}

### Diff-in-Diff Example II

\begin{example}
	In 1993 Georgia initiated a HOPE scholarship program to let state residents with at least a B average in high school attend public college in Georgia for free. Did it increase college enrollment?  
\begin{itemize}
		\item<2-> We can use a diff-in-diff model to measure the effect of HOPE scholarship on enrollments 
		\item<3-> Georgia and nearby states: if not for HOPE, changes should be the same over time
		\item<4-> Treatment period: 1992
		\item<5-> Treatment: Georgia 
	\end{itemize}
	\begin{align*}
	\only<6->{\Delta_i \Delta_t Enrolled &= (\text{GA}_{after}-\text{GA}_{before})-(\text{neighbors}_{after}-\text{neighbors}_{before})\\}
	\only<7->{\widehat{Enrolled_{it}} &= \beta_0+\beta_1 Georgia_{it}+\beta_2 After_{it}+\beta_3 Georgia_{it} * After_{it}\\}
	\end{align*}
\end{example}

### Diff-in-Diff Example III

```{r, echo=F}
HOPE<-read.csv("../Data/HOPE.csv")
```

\columnsbegin

\column{.5\textwidth}

\tiny 

```{r}
DND<-lm(InCollege~Georgia+After+AfterGeorgia, data=HOPE)
summary(DND)
```

\column{.5\textwidth}

\scriptsize 

$$\widehat{Enrolled_{it}}=0.406-0.105 Georgia_{i}-0.004 After_{t}+0.089 Georgia_{i} * After_{t}$$

\columnsend

### Diff-in-Diff Example: Interpreting Coefficients

$$\widehat{Enrolled_{it}}=0.406-0.105 Georgia_{i}-0.004 After_{t}+0.089 Georgia_{i} * After_{t}$$

- $\beta_0$: A **non-Georgian** **before** 1992 was 40.6\% likely to be a college student
- $\beta_1$: **Georgians** **before** 1992 were 10.5\% less likely to be college students than neighboring states
- $\beta_2$: **After** 1992, **non-Georgians** are 0.4\% less likely to be college students
- $\beta_3$: **After** 1992, **Georgians** are 8.9\% more likely to enroll in colleges than neighboring states
- \alert{Treatment effect: HOPE increased enrollment likelihood by 8.9\%}

### Diff-in-Diff Example: Comparing Group Means 

$$\widehat{Enrolled_{it}}=0.406-0.105 Georgia_{i}-0.004 After_{t}+0.089 Georgia_{i} * After_{t}$$

- A group mean for a dummy $Y$ is $E[Y=1]$, i.e. the probability a student is enrolled:
    - **Non-Georgian enrollment probability pre-1992**: $\beta_0=0.406$
    - **Georgian enrollment probability pre-1992**: $\beta_0+\beta_1=0.406-0.105=0.301$
    - **Non-Georgian enrollment probability post-1992**: $\beta_0+\beta_2=0.406-0.004=0.402$
    - **Georgian enrollment probability post-1992**: $\beta_0+\beta_1+\beta_2+\beta_3=0.406-0.105-0.004+0.089=0.386$

### Diff-in-Diff Example: Comparing Group Means in `R`

\columnsbegin

\column{.5\textwidth}

\scriptsize 
```{r}
# group mean for non-Georgian before 1992
HOPE %>%
  filter(Georgia==0 & After==0) %>%
  summarize(prob=mean(InCollege))

# group mean for Georgian before 1992
HOPE %>%
  filter(Georgia==1 & After==0) %>%
  summarize(prob=mean(InCollege))
```

\column{.5\textwidth}

\scriptsize 

```{r}
# group mean for non-Georgian AFTER 1992
HOPE %>%
  filter(Georgia==0 & After==1) %>%
  summarize(prob=mean(InCollege))

# group mean for Georgian AFTER 1992
HOPE %>%
  filter(Georgia==1 & After==1) %>%
  summarize(prob=mean(InCollege))
```

\columnsend

### Diff-in-Diff Example Summary

$$\widehat{Enrolled_{it}}=0.406-0.105 Georgia_{it}-0.004 After_{it}+0.089 Georgia_{it} * After_{it}$$

\begin{table}\small 
		\centering 
		\begin{tabular}{rrrr}
		 & Neighbors & Georgia & \textcolor{red}{Group Diff. ($\Delta Y_i$)}\\ \toprule 
		 Before & $0.406$ & $0.301$ & \textcolor{red}{$-0.105$}\\
		 After & $0.402$ & $0.386$ & \textcolor{red}{$0.016$}\\ \midrule 
		 \textcolor{blue}{Time Diff. ($\Delta Y_t$)} & \textcolor{blue}{$-0.004$}	& \textcolor{blue}{$0.085$} & \textcolor{purple}{$0.089$}\\ \bottomrule 
		 & & & \textcolor{purple}{Diff-in-diff ($\Delta \Delta Y$)}\\ 
		\end{tabular}
	\end{table}

\onslide<2->\begin{align*}
	\Delta_i \Delta_t Enrolled &= (\text{GA}_{after}-\text{GA}_{before})-(\text{neighbors}_{after}-\text{neighbors}_{before})\\
	&=(0.386-0.301)-(0.402-0.406)\\
	&=(\text{\textcolor{blue}{0.085}})-(-\text{\textcolor{blue}{0.004}})\\
	&=\text{\textcolor{purple}{0.089}}\\
	\end{align*}

### Diff-in-Diff Time Graph

```{r, echo=F}
HOPE %>%
  ggplot(data=., aes(x=After, y=InCollege, color=factor(Georgia)))+
  geom_smooth(method="lm", se=FALSE)+
  #scale_y_continuous(breaks=c(0,1,0.125), limits=c(0.3,0.5))+
  scale_x_discrete(breaks=seq(0,1,1), labels=c("Before", "After"), name="Before or After HOPE")+
  guides(color=guide_legend(title="Georgia"))+ylab("Probability of Being Enrolled in College")
```

```{r, experiment, eval=F, echo=F}
HOPE %>%
  mutate(After=as.factor(After)) %>% 
  ggplot(data=., aes(x=factor(After), y=InCollege), color=factor(Georgia))+
  geom_smooth(method="lm", se=FALSE)+
  scale_y_continuous(breaks=c(0,1,0.125), limits=c(0.3,0.5))+
  scale_x_discrete(breaks=seq(0,1,1), labels=c("Before", "After"), name="Before or After HOPE")+
  guides(color=guide_legend(title="Georgia"))+ylab("Probability of Being Enrolled in College")
```

```{r, experiment1, eval=F, echo=F}
HOPE %>%
  do(lm(InCollege~Georgia+After+AfterGeorgia, data=.) %>%
  predict %>%
  data_frame(Pred=.) %>%
  bind_cols(HOPE))
  ggplot(data=HOPE, aes(Year, y=Pred))+
  geom_line(aes(color=factor(Georgia)))
  #scale_y_continuous(breaks=c(0,1,0.125), limits=c(0.3,0.5))+
  #scale_x_discrete(breaks=seq(0,1,1), labels=c("Before", "After"), name="Before or After HOPE")+
  #guides(color=guide_legend(title="Georgia"))+ylab("Probability of Being Enrolled in College")

```


# Generalizing DND Models

### Generalizing DND Models

- DND can be **generalized** with a two-way fixed effects model:  	
$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 (\text{Treated}_i * \text{After}_{it})+\nu_{it}$$
    - $\alpha_i$: **group fixed effects** (treatments/control groups)
    - $\theta_t$: **time fixed effects** (pre/post treatment)
- Allows *many* periods, and treatment(s) can occur at different times to different units (so long as some do not get treated)
- Can also add control variables that vary within units and over time 
$$\widehat{Y_{it}}=\alpha_i+\theta_t+\beta_3 (\text{Treated}_i * \text{After}_{it})+\beta_4 X_{it}+\nu_{it}$$

### The Georgia HOPE Example, Generalized

\begin{example}
$$\widehat{Enrolled_{it}} = \alpha_i+\theta_t+\beta_3 Georgia_{it} * After_{it}$$
\end{example}

### The Georgia HOPE Example, Generalized II

\columnsbegin

\column{.45\textwidth}

\scriptsize

- `StateCode` is a variable for the State $\implies$ create State fixed effect
- `Year` is a variable for the year $\implies$ create year fixed effect
- Using LSDV method (note we must ensure both `StateCode` and `Year` are `factor` variables!): 

\column{.55\textwidth}

\tiny 

\onslide<4-> 

```{r}
DND_fe<-lm(InCollege~AfterGeorgia+factor(StateCode)+factor(Year), data=HOPE)
summary(DND_fe)
```

\columnsend

### Intuition Behind DND Models

- Diff-in-diff models are the quintessential example of exploiting \alert{natural experiments} 
    - A major change at a point in time (change in law, a natural disaster, political crisis) separates groups where one is affected and another is not---identifies the effect of the change (treatment) 
- One of the cleanest and clearest causal **identification strategies**

\begin{figure}
		\includegraphics[height=2in]{rct}
	\end{figure}

# Example: "The" Card-Kreuger Minimum Wage Study

### DND Example: Card & Kreuger (1994)

\begin{example}
	\emph{The} controversial minimum wage study, Card \& Kreuger (1994) is a quintessential (and clever) diff-in-diff.\\ \medskip  
	
	\small Card, David, Krueger, Alan B, (1994), ``Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania,'' \emph{American Economic Review} 84 (4): 772–793	
\end{example}

### Minimum Wage

\begin{columns}
		\begin{column}[c]{8cm}
			\begin{itemize}
					\item Economic theory: increases in minimum wage ($\bar{W}$) move us up a downward-sloping demand curve for labor 
					\item A surplus of labor: disemployment 
			\end{itemize}
		\end{column}
		\begin{column}[c]{7cm}
			\begin{tikzpicture}[y=.5cm, x=.5cm]\scriptsize 
				 \draw[->] (0,0) -- (11,0) coordinate (x axis) node[right]{L};
 				\draw[->] (0,0) -- (0,11) coordinate (y axis) node[above]{$W (\$)$};
				\draw[very thick,blue] (0,10) -- (10,0) node[right] {D};
				\draw[very thick,red] (0,0) -- (10,10) node[right] {S};	
				\draw[very thick, dotted, black] (0,5) node[left] {W*}-- (5,5) -- (5,0) node[below] {L*};
				\draw[very thick, black] (0,7) node[left]{$\bar{W}$} -- (10,7);
				\draw[very thick,black, dashed] (3,0) node[below]{\textcolor{blue}{$\bar{L_D}$}}-- (3,7);
				\draw[very thick,black, dashed] (7,0) node[below]{\textcolor{red}{$\bar{L_S}$}}-- (7,7);
				\draw[<->, very thick, black] (3, 1.5)--node[above]{Surplus}(7,1.5);
				\draw[ultra thick, ->] (-1.5, 5)--(-1.5,7);
 			\end{tikzpicture}
		\end{column}
	\end{columns}

### Card & Kreuger (1994): Background

- Card & Kreuger (1994)	compare employment in fast food restaurants on New Jersey and Pennsylvania sides of border between February and November 1992.
- Pennsylvania & New Jersey both had a minimum wage of \$4.25 before February 1992
- In February 1992, New Jersey raised minimum wage from \$4.25 to \$5.05

\begin{figure}
		\includegraphics[height=1.5in]{PANJ}	
\end{figure}

### Card & Kreuger (1994): Background II

- If we look only at New Jersey before & after change: 
    - **Omitted variable bias**: macroeconomic variables (there's a recession!), weather, etc.
		- Including PA as a control will control for these time-varying effects if they are national trends
- Surveyed 400 fast food restaurants on each side of the border, before & after min wage increase
    - \alert{Key assumption}: Pennsylvania and New Jersey follow parallel trends, 
		- **Counterfactual**: if not for the minimum wage increase, NJ employment would have changed similar to PA employment

\begin{figure}
		\includegraphics[height=1.5in]{PANJ}	
\end{figure}

### Card & Kreuger (1994): Comparisons

\begin{figure}
		\includegraphics[height=3in]{CardKreuger1}
\end{figure}

### Card & Kreuger (1994): Summary

\begin{figure}
		\includegraphics[height=3in]{CardKreuger2}
\end{figure}

### Card & Kreuger (1994): Summary II 

\begin{figure}
		\includegraphics[height=3in]{CardKreuger3}
\end{figure}

### Card & Kreuger (1994): Identification Strategy and Model 

$$\widehat{Employment_{i t}}=\beta_0+\beta_1 NJ_	{i}+\beta_2 After_{t}+\beta_3(NJ_i * After_t)$$

- PA Before: $\beta_0$
- PA After: $\beta_0+\beta_2$
- NJ Before: $\beta_0+\beta_1$
- NJ After: $\beta_0+\beta_1+\beta_2+\beta_3$
- **Diff-in-diff**: $(NJ_{after}-NJ_{before})-(PA_{after}-PA_{before})$

\begin{table}\small 
		\centering 
		\begin{tabular}{rrrr}
		 & PA & NJ & \textcolor{red}{State Diff. ($\Delta Y_i$)}\\ \toprule 
		 Before & $\beta_0$ & $\beta_0+\beta_1$ & \textcolor{red}{$\beta_1$}\\
		 After & $\beta_0+\beta_2$ & $\beta_0+\beta_1+\beta_2+\beta_3$ & \textcolor{red}{$\beta_1+\beta_3$}\\ \midrule 
		 \textcolor{blue}{Time Diff. ($\Delta Y_t$)} & \textcolor{blue}{$\beta_2$}	 & \textcolor{blue}{$\beta_2+\beta_3$} & \textcolor{purple}{$\beta_3$}\\ \bottomrule 
		 & & & \textcolor{purple}{Diff-in-diff ($\Delta \Delta Y$)}\\ 
		\end{tabular}
\end{table}

### Card & Kreuger (1994): Comparisons

\begin{figure}
		\includegraphics[height=2in]{CardKreuger5.png}
\end{figure}


