---
title: "Lecture 12: Model Specification Strategies"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "October 31, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(ggplot2) 
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, dev.args=list(bg="transparent"))
```

```{r, echo=FALSE}
# load data set and run regression again 

library("foreign") #for importing .dta files
CASchool<-read.dta("../Data/caschool.dta")

# run miltivariate regression again 
multireg<-lm(testscr~str+el_pct, data=CASchool)
```

# Model Specification

### Model Specification

- The big challenge in applied econometrics is choosing how to \alert{specify a model} to regress
- Every dataset is different, every study has a different goal
    - there is no bright line rule, only a set of guidelines and skills that you can only learn by doing! 
- But here are some helpful tips and frequent problems (and solutions)	

### Model Specification: Process

1. **Identify your question of interest:** what do you want to know? What marginal effect(s) do you want to estimate?
2. **Think about possible sources of endogeneity:** what *other* variables would cause  \alert{omitted variable bias} if we left them out? Can we get data on them too? 
    - Again: must BOTH (1) affect $Y$ AND (2) be correlated with $X$
    - This requires much of your economic intuitions: $R^2$ and statistical measures cannot tell you everything!
3. **Run multiple models and check the robustness of your results:** does the size (or direction) of your marginal effect(s) of interest change as you change your model (i.e. add more variables)?
4. **Interpret your results**
    - Are they statistically significant? 
    - Regardless of *statistical* significance, are they **economically meaningful**? 
        - Why should we care? 
        - How big is "big"? 

### Proxy Variables

- Ideally, we would want a randomized control experiment to assign individuals to treatment
- But with observational data, $\epsilon_i$ depends on other factors
    - *If* we can observe and measure these factors, then include them in the regression
- If we can't directly measure them, often we can include variables *correlated* with these variables to \alert{proxy} for the effects of them!

### Proxy Variables: Example

\begin{example}
	Consider test scores and class sizes again. What about learning opportunities outside of school? 	
	\begin{itemize}[<+->]
		\item Probably a bias-causing omitted variable (affects test score and correlated with class size) but we can't measure it! 
		\item But suppose we \emph{can} measure a variable $V$, and significantly, $corr(V,Z)\neq 0$	
		\item e.g. we have data on the percent of students who get a free or subsidized lunch (`meal\_pct`)
		\item This is a good \textbf{proxy} for income-determined learning opportunities outside of school
		\begin{itemize}
			\item \%meal is correlated with Income, which is correlated with both class size and test score
			\item So this is a good \emph{indirect} measure of Income 
		\end{itemize} 
	\end{itemize}
\end{example}

### Proxy Variables: Example II 

- We've been assuming we don't have data on average district income, we would expect `meal_pct` to be strongly negatively correlated with income
- Just kidding, we do have data on `avginc`, but we'll only use it to confirm our suspicion: 

```{r, echo=FALSE, results="asis"}
CAproxy<-subset(CASchool, select=c("testscr", "str", "el_pct", "avginc","meal_pct"))

corrproxy<-cor(CAproxy)

library("stargazer")
stargazer(corrproxy, type="latex", header=FALSE, float=FALSE)
```

- We can see `meal_pct` is strongly (negatively) correlated with income, as expected

### Proxy Variables: Example III 

\columnsbegin
\column{.45\textwidth}

\tiny 
```{r}
proxyreg<-lm(testscr~str+el_pct+meal_pct, data=CAproxy)
summary(proxyreg)
```

\column{.55\textwidth}


\begin{table} 
	\centering 
				\setlength{\tabcolsep}{0cm}
	\onslide<2->\begin{tabular}{ccccccccccc}\small 
	  $\widehat{\text{Test Score}}=$ & 700.15& $-$ & $1.00$ & $STR$ & $-$ & $0.122$ & $el\_pct$ & $-$ &$0.547$ & $meal\_pct$\\
	& \scriptsize (4.68) & & \scriptsize (0.24) & & & \scriptsize (0.03) & & & \scriptsize (0.02) & \\
	\end{tabular}
\end{table} 

\columnsend

### Interpretting Control Variables 

\begin{table} 
	\centering 
				\setlength{\tabcolsep}{0cm}
	\onslide<2->\begin{tabular}{ccccccccccc}\small 
	  $\widehat{\text{Test Score}}=$ & 700.15& $-$ & $1.00$ & $STR$ & $-$ & $0.122$ & $el\_pct$ & $-$ &$0.547$ & $meal\_pct$\\
	& \scriptsize (4.68) & & \scriptsize (0.24) & & & \scriptsize (0.03) & & & \scriptsize (0.02) & \\
	\end{tabular}
\end{table} 

- Is `meal_pct` causal? 
- Getting rid of programs in districts where a large percentage of students need them would boost test scores A LOT! (So probably not causal...) 
- `meal_pct` likely correlated with other things in $\epsilon$ (like outside learning opportunities!). 
    - In fact, that's *exactly why* we included it as a variable! 
- We don't need the OLS estimate on `meal_pct` to be unbiased! 
    - We only care about getting the estimate on `str` to be unbiased! 

### Control Variables

- A \alert{control variable} is a regressor variable **note** of interest, but included to hold factors constant that, if omitted, would bias the causal effect of interest
- The control variable may still be correlated with omitted causal factors in $\epsilon$
    - Estimators ($\hat{\beta}$'s) on control variables can be biased and that is OK!
    - So long as we have unbiased estimators ($\hat{\beta}$'s) on the regressors we do care about!
- Control variables allow us to proceed \emph{as if} $X$ were randomly assigned 

### Model Specification Warning

\begin{alertblock}{Do NOT just try to maximize $R^2$ or $\bar{R}^2$}
		\begin{itemize}[<+->]
			\item A high $R^2$ or $\bar{R}^2$ means that the regressors explain variation in $Y$
			\item A high $R^2$ or $\bar{R}^2$ does \emph{NOT} mean you have eliminated omitted variable bias
			\item A high $R^2$ or $\bar{R}^2$ does \emph{NOT} mean you have an unbiased estimate of a causal effect
			\item A high $R^2$ or $\bar{R}^2$ does \emph{NOT} mean included variables are statistically significant
		\end{itemize}
\end{alertblock}

# Applications of Simple Multivariate OLS Models 

## Sacerdote (2004) on Peer Effects

### Sacerdote (2004) on Peer Effects

Sacerdote, Bruce, (2004), "Peer Effects with Random Assignment: Results from Dartmouth Roommates" *Quarterly Journal of Economics* 116(2):681-704

\begin{figure}
		\includegraphics[height=2in]{universitypeers}	
\end{figure}

### Paper Motivation

- What determines student outcomes in college? (GPAs, fraternity enrollment, alcohol/drug use, etc) 
- Effects of peer groups

\begin{figure}
		\includegraphics[height=2in]{universitypeers}	
\end{figure}

### Paper Motivation II

- "Standard" way to estimate peer effects: regress student $i$'s outcomes/behavior on *other students'* outcomes/behavior

$$GPA_i = \beta_0+\beta_1 OwnBehavior_i + \beta_2 RoommateBehavior_i + \epsilon_i $$

- Problems with this approach: 
    1. Individuals **self-select** into peer groups
    2. If two roommates $A$ and $B$ influence each other, how do we disentangle causal effect of $B \rightarrow A$ vs. $A \rightarrow B$?
    3. Are peer effects actually driven by students' own backgrounds, or by their actual choices? 

### Paper Motivation III

\begin{align*}
			corr(OwnBehavior, \epsilon) &\neq 0	\\
			corr(RoomateBehavior, \epsilon)& \neq 0 \\
			E[\epsilon|OwnBehavior,RoommateBehavior] & \neq 0 	\\
\end{align*}

\begin{figure}
		\includegraphics[height=2in]{universitypeers}	
\end{figure}

### Sacerdote's Identification Strategy

- Freshmen entering Dartmouth College are **randomly** assigned to dorms & roommates
- Removes self-selection of peer groups by shared characteristics	
- Random assignment: roommate A's background characteristics are uncorrelated with roommate B's background characteristics

\begin{figure}
		\includegraphics[height=2in]{universitypeers}	
\end{figure}

### The Relevant Institutions

- Freshmen entering Dartmouth College are randomly assigned to dorms & roommates
- Each incoming freshman fills out a questionnaire:
    1. I smoke. (Y/N)
    2. I like to listen to music while studying. (Y/N)
    3. I keep late hours. (Y/N)
    4. I am more neat than messy. (Y/N)
    5. I am (Male/Female). 	
- There are $2^5=32$ combinatorial possibilities of answers to the questions
- Students are assigned to roommates/dorms **at random**, conditional on their combination of answers to the 5 survey answers

### The Data

- Data from Dartmouth's database of students: history of dorm assignments \& term-by-term academic performance
- Data on pre-treatment characteristics (SAT scores, high school class rank, private/public HS, home state, academic index)
- Outcome variables: GPA, time to graduation, frat membership, major choice, participation in athletics
- Survey of Incoming Freshman: if student drank beer in last year and expectation of graduating with honors
- Sample of 1589 students	
- Create dummy variable for each block to control for covariates (we'll talk later about \alert{dummy variables} and \alert{fixed effects} like this)

### The Data: Summary Statistics

\begin{figure}
		\includegraphics[height=3in]{sacerdote1}	
\end{figure}

### The Data: Summary Statistics II 

\begin{figure}
		\includegraphics[height=3in]{sacerdote2}	
\end{figure}

### The Data: Demonstrating Random Assignment

\begin{figure}
		\includegraphics[height=3in]{sacerdote3}	
\end{figure}

### The Data: Demonstrating Random Assignment II 

\begin{figure}
		\includegraphics[height=3in]{sacerdote3c}	
\end{figure}

### Sacerdote's Empirical Model (Basically)

$$GPA_i=\beta_0+\beta_1ACA_i+\beta_2ACA_j+\epsilon_i$$

- Student $i$ and roommate $j$
- $ACA$: Index of academic performance (broken down into different metrics)  
- Other outcomes of interest (besides $GPA$: graduation, major, fraternity, athlete

### Regression Results

\begin{figure}
		\includegraphics[height=2in]{sacerdote5}	
\end{figure}
\scriptsize 

- For every 1 point increase (decrease) in the roommate's GPA, a student's GPA increased (decreased) about .12 points
- If you would have been a 3.0 student with a 3.0 roommate, but you were assigned to a 2.0 roommate, your GPA would be 2.88

### Regression Results: Academic Performance

\begin{figure}
		\includegraphics[height=2in]{sacerdote6}	
	\end{figure}

### Regression Results: Social Outcomes 

\begin{figure}
		\includegraphics[height=3in]{sacerdote9}	
\end{figure}

### Sacerdote's Conclusions

- Peer effects are very strong!
- Important influences in Freshman year performance (GPA) and activities (joining a social organization)
- Not important for choosing a major

\begin{figure}
		\includegraphics[height=2in]{universitypeers}	
\end{figure}


## Duggan and Levitt (2002) on Corruption in Sumo 

### Duggan and Levitt (2002) on Corruption in Sumo 

Duggan, Mark and Steven D. Levitt (2002), "Winning Isn't Everything: Corruption in Sumo Wrestling," *American Economic Review* 92(5): 1594-1605

\begin{figure}
		\includegraphics[height=2in]{sumo}	
\end{figure}

### Paper Motivation

- How can we *understand* and *detect* corruption?
- A very important and consequential economic problem, nearly impossible to measure

\begin{figure}
		\includegraphics[height=2in]{rentseeking}	
\end{figure}

### Some Background: CPI 

- **Transparency International: Corruption Perceptions Index (CPI)**
    - Index of how people \emph{perceive} corruption in their country
    -168 countries
    - Draws on a composite of 4 expert opinion surveys
    - Scale from 0-100, 0: high levels of perceived corruption; 100: low levels of perceived corruption

### Some Background: CPI II 

\begin{figure}
		\includegraphics[height=2.5in]{tcpi2015}
		\caption*{\href{http://www.transparency.org/cpi2015}{Transparency International CPI 2015}}	
\end{figure}

### Some Background: CPI III 

\begin{figure}
		\includegraphics[height=2in]{tcpi1}
		\caption*{\href{http://www.transparency.org/cpi2015}{Transparency International CPI 2015}}	
\end{figure}

### Some Background: CPI IV

\begin{figure}
		\includegraphics[height=1.75in]{tcpi2}
		\caption*{\href{http://www.transparency.org/cpi2015}{Transparency International CPI 2015}}	
\end{figure}

### Duggan and Levitt (2002): Relevant Institutions 

- 2000 year history, national sport of Japan, extremely ritualistic
- Japan is a country with low corruption (CPI: 75, Rank 18th best)
- Good data available
- Situation is ripe for cheating! So when/why does it happen? 

\begin{figure}
		\includegraphics[height=2in]{sumo}	
\end{figure}

### Duggan and Levitt (2002): Relevant Institutions II

- Tournaments (*basho*), 66 wrestlers (*rikishi*), 15 bouts each
- Wrestlers with 8+ wins (*kachi-koshi*) *move up* in rankings (*banzuke*)
- Those with a losing record ($<$8 wins) (*maki-koshi*) *fall* in rankings 

\begin{figure}
		\includegraphics[height=2in]{sumo1}	
\end{figure}

### Duggan and Levitt (2002): Relevant Institutions III

- A marginal win generates a 2.5 rank increase
- But movement from 7 to 8 wins produces almost an 11 rank increase!
- Rank signals prestige, moving up a single rank is worth about $3,000/year 
- Top 5th-10th ranked wrestlers make $250,000/year  

\begin{figure}
		\includegraphics[height=2in]{sumo1}	
\end{figure}

### Duggan and Levitt (2002): Relevant Institutions IV

- Consider 2 wrestlers: \textcolor{teal}{A (8-6)} vs \textcolor{magenta}{B (7-7)} going into final (15th) match
- Return to winning for \textcolor{magenta}{B (7-7)} is much higher than for \textcolor{teal}{A (8-6)}
- \textcolor{teal}{A (8-6)} throws the match to \textcolor{magenta}{B (7-7)}, who must return the favor in later tournaments if \textcolor{teal}{A} finds himself in the same 7-7 position

\begin{figure}
		\includegraphics[height=2in]{sumo}	
\end{figure}

### Duggan and Levitt (2002): Data

- All official top-rank sumo matches from January 1989-January 2000
- Six tournaments per year, nearly 70 wrestlers per tournament		
- Tournaments last 15 days with one match per wrestler 
- 64,000 wrestler-matches

### The Theoretical vs. Actual Probability of Winning

\begin{figure}
		\includegraphics[height=2in]{sumo2}	
\end{figure}

- Theoretical (binomial) probability of winning 8 times: 19.6%
- Actual probability (from data): 26%
- Much higher probability for 8 wins than it should be! (& lower for 7)
    - Could this be from rampant cheating!?

### The Model 

\begin{figure}
		\includegraphics[height=0.5in]{sumo3}	
\end{figure}

- $Win=1$ if wrestler $i$ beats wrestler $j$ in tournament $t$ on day $d$
- Bubble =1 if wrestler ($i$) is on margin (7-7), -1 if opponent ($j$) is on margin, =0 if neither are on margin 
- $Rankdiff$: difference in rank between wrestlers 
- Wrestler $\lambda$ and tournament $\delta$ fixed effects

### Initial Results 

\begin{figure}
		\includegraphics[height=2.7in]{sumo4}	
\end{figure}

\scriptsize

- Frequency of rigging increases as tournament nears end (day 15) 
- On day 15, 7-7 wrestlers on margin win 25% more often than they otherwise should 

### Interpretation of Initial Results

- Two alternative hypotheses to explain results:
    1. Match rigging (corruption)
    2. Effort: wrestlers on margin (7-7) just fight harder! Wrestlers with 8 wins are more complacent (already made it) 
- To test, look for evidence of **reciprocity** agreements over time 
    - If these tacit agreements to rig matches exist, wrestlers from stable A should have very high win rates when on the margin against wrestlers from stable B, and vice versa 
    
### Interaction Effects

\begin{figure}
		\includegraphics[height=2.7in]{sumo5}	
\end{figure}
\scriptsize

- Last row (before $R^2$): wrestler's success strongly increases with overall success rates of playing wrestlers on the bubble from other stables
- For each 10% increase in success in other bubble matches between these two stables, the wrestler on the bubble is 2.7% more likely to win

### Reciprocity

\begin{figure}
		\includegraphics[height=2.7in]{sumo6}	
\end{figure}
\scriptsize

-Wrestlers who won a bubble match previously tend to do worse when playing same opponent - throw the match to them in reciprocity! 

## Fisman and Miguel (2007) on U.N. Parking Tickets

### Fisman and Miguel (2007) on U.N. Parking Tickets

Fisman, Raymond and Edward Miguel, (2007), "Corruption, Norms, and Legal Enforcement: Evidence from Diplomatic Parking Tickets," *Journal of Political Economy* 115(6): 1020-1048

\begin{figure}
		\includegraphics[height=2in]{parkingticket}	
\end{figure}

### Fisman and Miguel (2007): Motivation

- What determines the level of corruption?
    1. Legal environment of country
    2. Culture or social norms 
- How to identify the true source(s)?

\begin{figure}
		\includegraphics[height=2in]{rentseeking}	
\end{figure}

### Fisman and Miguel (2007): Motivation II

- U.N. Diplomats are given immunity from prosecution or lawsuits in the U.S.
- Reciprocal agreements with other countries, designed to protect diplomats in unfriendly environments
- Diplomatic license plates in NYC are identified, get ticket, but no way to enforce
- "The best free parking pass in town"

\begin{figure}
		\includegraphics[height=1in]{parkingticket}	
		\includegraphics[height=1in]{diplomatplates}	
\end{figure}

### Fisman and Miguel (2007): The Data

- Between 11/1987 and 12/2002, 150,000 unpaid parking tickets, fines totaling $18,000,000
- 30 Days to pay a fine, afterwards a 110% penalty. After 70 days, recorded as unpaid violation
- Individual violation-level data: license plate, name, country of origin, date & time of violation, fine, amount paid (if any)
- 43% were violations of "no standing/loading zone"
- 20% of cases, the car was registered to the diplomatic mission (not personal)
- Scale fines by the size of the country's mission 

\begin{figure}
		\includegraphics[height=1in]{parkingticket}	
		\includegraphics[height=1in]{diplomatplates}	
\end{figure}

### Fisman and Miguel (2007): Identification Strategy

- Becker's (1968) rational crime model says with no punishment$\implies$ rational for all diplomats to never pay parking fines
- But large variation in data! Unpaid fines are strongly correlated with country's score on corruption index! 
- Home country corruption norms are an important predictor of diplomats breaking the law
    - Low corruption countries' diplomats tend to pay the fine even where they are not legally compelled to
    - High corruption countries' diplomats rack up massive fines

### Fisman and Miguel (2007): Identification Strategy II 

- **Natural experiment**: post-9/11, NYC began cracking down on enforcement
- Diplomats with 3+ unpaid parking tickets had diplomat plates revoked
- Led to immediate 98% reduction in unpaid parking tickets
- So enforcement matters as well as corruption norms

\begin{figure}
		\includegraphics[height=2in]{parking1}	
\end{figure}

### Fisman and Miguel (2007): Data 

\begin{figure}
		\includegraphics[height=2.5in]{parking2}	
\end{figure}

### Fisman and Miguel (2007): Data II

\begin{figure}
		\includegraphics[height=2.5in]{parking3}	
\end{figure}

### Fisman and Miguel (2007): Empirical Model

$$\text{Unpaid Violations}=\beta_0+\beta_1\text{Corruption}+\beta_2\text{Enforcement}+\beta_3\text{Diplomats}+...+\beta_k\text{Controls}$$
	
\begin{figure}
		\includegraphics[height=1in]{parkingticket}	
		\includegraphics[height=1in]{diplomatplates}	
\end{figure}

### Fisman and Miguel (2007): Results 

\begin{figure}
		\includegraphics[height=2.5in]{parking4}	
\end{figure}

### Fisman and Miguel (2007): Results II

\begin{figure}
		\includegraphics[height=2.5in]{parking7}	
\end{figure}

### Fisman and Miguel (2007): Results III

\begin{figure}
		\includegraphics[height=2.5in]{parking5}	
\end{figure}

