---
title: "Lecture 15: Better Data Wrangling with `dplyr`"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "November 19, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(tidyverse)
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, dev.args=list(bg="transparent"))
```

### Some Meta Thoughts About Code

- There is a tradeoff between **precision** and **concision** in coding:
    - `subset(diamonds, x == 0 & y == 0 & z == 0)` 
    - vs. 
    - `diamonds[diamonds$x == 0 & diamonds$y == 0 & diamonds$z == 0, ]`
- It would be ideal for code to be "self-documenting" and easily readable to observers without excess explanation

### Some Meta Thoughts About Code II 

- Compare the following commands, which both subset the `gapminder` data to look only at year and life expectancy for Cambodia

```{r, eval=FALSE}
gapminder[gapminder$country=="Cambodia", c("year", "lifeExp0")]
```

```{r, eval=FALSE}
gapminder %>%
  filter(country == "Cambodia") %>%
  select(year, lifeExp)
```

- Which is more intuitive to read and understand what we're doing? (without comments!)
    - The first is using Base `R`, the second uses `dplyr`
    
### The Pipe Operator, `%>%`

- The \alert{"pipe" operator}, `%>%` (from package `magrittr`) will change your coding life 
- Keyboard shortcut in R Studio:
    - `CTRL+Shift+M` (Windows)
    - `Cmd+Shift+M` (Mac)
- `%>%` "pipes" the *output* of everything to the *left* of the pipe into the *input* on right
- Running some function `f` on object `x` as `f(x)` can be "piped" as `x %>% f`
    - i.e. "take `x` and then perform function `f` on it"

\begin{figure}
  \includegraphics[height=1.4in]{magrittr}
\end{figure}

### The Pipe Operator, `%>%` II

- With ordinary math functions, we read operations from outside$\leftarrow$(inside): 
$$g(f(x))$$
i.e. take `x` and then perform function `f` on `x`, then perform function `g` on that result
- With pipes, we read operations from left$\rightarrow$ right: 

\onslide<3->
```{r, eval=FALSE}
x %>% f %>% g
``` 

\onslide<3->
take `x` and then perform function `f` on it, then perform function `g` on that result 

\onslide<4->
- So read `%>%` mentally as "and then"

### The Pipe Operator, %>% Example

```{r, echo=FALSE}
library("gapminder")
```

\columnsbegin

\column{.5\textwidth}

\tiny 
```{r}
# look at top 6 rows
head(gapminder) 
```

\column{.5\textwidth}

\tiny
```{r}
# use the pipe instead 
gapminder %>% head()
```

\columnsend

### The `tidyverse`

> "The tidyverse is an opinionated  collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures. 

- Core packages include ones we've discussed before: `ggplot2`, `dplyr`, `magrittr` among several others (`tidy` `readr`, `purr`, `forcats`, `stringr`)
- Loading any `tidyverse` package loads `magrittr` (so you can use `%>%`)
- Learn more at [tidyverse.org](tidyverse.org) 

\begin{figure}
  \includegraphics[height=1.45in]{tidyverse}
\end{figure}

### The `tidyverse` II 

- Easiest to just load the core tidyverse all at once
    - Note loading the tidyverse is "noisy", it will spew a lot of messages
    - Hide them with `suppressPackageStartupMessages()` and insert the `library()` command inside the `()` 

```{r, echo=FALSE}
detach("package:tidyverse", TRUE) # just to show start up message of tidyverse, let's unload it first
```

```{r, message=TRUE, warning=TRUE}
library("tidyverse")
```

### The `tidyverse` III

\begin{figure}
  \includegraphics[height=2in]{tidyverse2}
\end{figure}

### `dplyr` Verbs

- Base `R` is about running functions on nouns, e.g. `function(object)` 
- `dplyr` is all about using active English-language verbs to accomplish tasks

\onslide<3->

| Function | Does | 
|----------|------|
| `filter` | Keep only selected *observations* | 
| `select` | Keep only selected *variables* |
| `arrange` | Reorder rows (e.g. in numerical order) |
| `mutate` | Create new variables | 
| `recode` | Change a variable's values or categories/`factor` levels |
| `summarize` | Collapse data into summary statistics| 
| `group_by` | Perform any of the above functions by groups/categories | 

### `dplyr` Verbs II 

- Syntax of any `dplyr` function is the same: `dyplrfunction(dataframe, condition)`, which returns a `data.frame`
    - Or if you prefer to try out the pipe `%>%`:
    
\onslide<2->

\columnsbegin

\column{.5\textwidth}

\scriptsize 
```{r, eval=FALSE}
# without the pipe 
dplyrfunction(dataframe, condition)
```

\column{.5\textwidth}

\scriptsize 
```{r, eval=FALSE}
# with the pipe 
dataframe %>% condition()
```

\columnsend

### `filter`

- `filter` keeps only selected **observations**

\scriptsize 

\onslide<2->

\columnsbegin

\column{.5\textwidth}

\tiny 
```{r}
# look only at African observations
# syntax without the pipe 
filter(gapminder, continent=="Africa")
```

\column{.5\textwidth}

\onslide<3->

\tiny 
```{r}
# look only at African observations
# syntax with the pipe 
gapminder %>%
  filter(continent=="Africa")
```

\columnsend

### `dplyr` Allows You to Test Without Overwriting

- Great thing about `dplyr` is we don't necessarily need to store our results as objects until we're ready 
    - Won't overwrite an object inccorectly, e.g.: 
    - `gapminder %>% select(country == "United States")` does not overwrite `gapminder`. 
- You can still deliberately save (and overwrite) objects with the assignment operator:
    - `gapminder <- gapminder %>% select(country == "United States")` *would* overwrite gapminder with *just* the U.S. observations

### `filter` Multiple AND Conditions

- `filter` multiple conditions with commas (implicitly, having multiple "AND" conditions)

\scriptsize 

\onslide<2->

```{r}
# look only at observations that are in Europe AND in 1997 
gapminder %>%
  filter(continent=="Europe", year==1997)
```

### `filter` OR Conditions

- `filter` multiple alternative conditions with | ("OR")

\scriptsize 

\onslide<2->

```{r}
# look only at observations that are in Europe OR in 1997 
gapminder %>%
  filter(continent=="Europe" | year==1997)
```

### `filter` Other Useful Operators

- We can `filter` by membership `%in%` a particular set (represented by a vector)

\scriptsize 

\onslide<2->

```{r}
# look only at observations that are in the "set" of (Europe, Africa, Asia) 
gapminder %>%
  filter(continent %in% c("Europe", "Africa", "Asia"))
```

### Compare to Base `R`'s `subset()` function

\begin{figure}
  \includegraphics[height=2.5in]{rsubset}
\end{figure}

### `select` 

- `select` keeps only selected **variables** 

\scriptsize 

\onslide<2->

```{r}
# Only keep country, year, and population variables
gapminder %>%
  select(country, year, pop)
```

### `select` Helper Functions

- `select` has a lot of nice helper functions
- Type `?select` to get more information
- Some examples (where `string` is some text that you are searching for):

\onslide<3->

| Function | Description |
|----------|-------------|
| `starts_with("string")` | Variable name begins with `string` |
| `ends_with("string")` | Variable name ends with `string` | 
| `contains("string")` | Variable name contains `string` in it |

### `select` Example

\scriptsize 

\onslide<2->

```{r}
# Only keep gdpPercap and othervariables that start with "c"
gapminder %>%
  select(gdpPercap,starts_with("c"))
```

### `select` Neg

- `select` allows you to "negate" columns with a negative sign (-)
    - A way of keeping "everything but" certain variables

\scriptsize 

\onslide<4->

```{r}
# keep all variables EXCEPT pop
gapminder %>%
  select(-pop)
```

### `rename` 

- `rename` changes the name of a variable in the following format: `newname=oldname` 

\scriptsize 

\onslide<2->

```{r}
# Rename gdpPercap to just GDP
gapminder %>%
  rename(GDP=gdpPercap)
```

### `arrange`

- `arrange` orders the observations (rows) in some logical order
    - e.g. (reverse) alphabetical, (reverse) numerical, largest to smallest (smallest to largest)

\scriptsize 

\onslide<3->

```{r}
# Sort by lifeExp 
gapminder %>%
  arrange(lifeExp)
```

### `arrange()` II 

\begin{figure}
  \includegraphics[height=2.5in]{arrange1}
\end{figure}

### `arrange` III 

- use `desc()` for descending order

\scriptsize 

\onslide<2->

```{r}
# Sort by country name (reverse alphabetically)
gapminder %>%
  arrange(desc(country))
```

### `arrange()` IV

\begin{figure}
  \includegraphics[height=2.5in]{arrange2}
\end{figure}

### `mutate`

- `mutate` creates a new variable according to some operation on other variables
    - syntax: `new.variable.name=operation`

\scriptsize 

\onslide<3->

```{r}
# make a GDP variable by multiplying gdpPercap and pop
gapminder %>%
  mutate(gdp= gdpPercap * pop)
```

### `mutate()` II 

\begin{figure}
  \includegraphics[height=2.5in]{mutate1}
\end{figure}

### `mutate()` III

\begin{figure}
  \includegraphics[height=2.5in]{mutate2}
\end{figure}

### `mutate` Multiple Variables at Once

- Can create multiple new variables with the same command using commas

\scriptsize 

\onslide<2->

```{r}
gapminder %>%
  mutate(gdp= gdpPercap * pop,
         gdp.billions=gdp/1000000000)
```

### `summarize` 

- `summarize`\footnote{Also the non-U.S. English spelling `summarise` works. `dplyr` was written by a Kiwi after all!} calculates desired summary statistics of a variable
    - common summary statistics: `n()`, `mean()`, `sd()`, `min()`, `max()`
    - `summarize` outputs a new `data.frame`
    - Can give a name to the summary variable as if you are `mutate`-ing a new variable

\scriptsize 

\onslide<5->

\columnsbegin

\column{.5\textwidth}

\scriptsize 
```{r}
# get average life expectancy 
gapminder %>%
  summarize(mean(lifeExp))
```

\column{.5\textwidth}

\onslide<6->

\scriptsize 
```{r}
# same thing, but calling the new variable "avg.LE" 
gapminder %>%
  summarize(avg.LE=mean(lifeExp))
```

\columnsend


### `summarize` II

\begin{figure}
  \includegraphics[height=2.5in]{summarize2}
\end{figure}

### `summarize` III

\begin{figure}
  \includegraphics[height=2.5in]{summarize1}
\end{figure}

### `summarize` IV

- Can `summarize` multiple variables at once 

\scriptsize 

\onslide<2->

```{r}
# get average life expectancy, gdp per capita, and population 
gapminder %>%
  summarize(avg.LE=mean(lifeExp),
            avg.GDP=mean(gdpPercap),
            abg.pop=mean(pop))
```

### `summarize` Other Statistics

\scriptsize 

\onslide<2->

```{r}
# get count, mean, sd, min, and max life expectancy 
gapminder %>%
  summarize(count.LE=n(),
            avg.LE=mean(lifeExp),
            sd.LE=sd(lifeExp),
            min.LE=min(lifeExp),
            max.LE=max(lifeExp))
```

### `group_by`

- If we have `factor` variables (such as `continent`), we can run all of our `dplyr` verb commands by group
- First we define the groups as the `continent` 

\scriptsize 

\onslide<3->

```{r}
gapminder %>% 
  group_by(continent) %>% # first group by continent
  summarize(mean_life=mean(lifeExp), # get averages for life exp 
            mean_gdp=mean(gdpPercap), # and gdp per capita
            mean_pop=mean(pop)) # and population by continent
```

### `group_by` Without the Pipe

- Since there are several steps going on here, let's think about what this would look like without the %>% operator: 

\scriptsize 

\onslide<2->

```{r}
# first generate a variable for the groups 
by_continent<-group_by(gapminder, continent)

# then summarize the variable 
summarize(by_continent, mean_life=mean(lifeExp),
          mean_gdp=mean(gdpPercap),
          mean_pop=mean(pop))
```

### `group_by` Example II

\scriptsize 

\onslide<2->

```{r}
# track change in average Life Expectancy over time
gapminder %>% 
  group_by(year) %>%
  summarize(mean.LE=mean(lifeExp),
            mean.GPD=mean(gdpPercap))
```

### `group_by` and `summarize_at` to Combine Statistics

- Use `summarize_at` to summarize multiple variables with multiple summary statistics
- Syntax: `summarize_at(vars(var1, var2), funs(stat1, stat2))` where `var1` and `var2` are your variables and `stat1` and `stat2` are the summary statistics you'd like (e.g. mean, median, etc)

\scriptsize 

\onslide<3->

```{r}
# get summary statistics (mean, median, sd) for lifeExp and gdpPercap over time 
gapminder %>%
  group_by(year) %>%
  summarize_at(vars(lifeExp, gdpPercap), funs(mean, median, sd))

```

### `tally` 

- `tally` is shorthand for just getting the counts of observations by group (instead of `summarize` and `n()`)

\scriptsize 

\onslide<2->

```{r}
gapminder %>%
  group_by(continent) %>%
  tally
```

### Piping Across Packages

- The `tidyverse` uses the same grammar and design philosophy, so you can (almost always) pipe things across packages and functions
- Example: graph the change in average life expectancy by continent over time

\scriptsize 

\onslide<2->

```{r, fig.height=2.5}
gapminder %>% # start with gapminder data 
  group_by(year, continent) %>% # create groups of years and of continents 
  summarize(mean_life = mean(lifeExp)) %>% # get average life expectancy for each group 
  ggplot(aes(year, mean_life, color = continent))+ # plot this over time 
  geom_point() + geom_smooth(method="lm")
```

### `dplyr` Cheatsheet

As usual, there is a fantastic cheatsheet for `dplyr` via RStudio

\begin{figure}
  \includegraphics[height=2.45in]{dplyrcheatsheet}
  \caption*{\href{https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf}{RStudio's dplyr Cheatsheet}}
\end{figure}

