---
title: "Lecture 3: Data and Descriptive Statistics"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "September 5, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(ggplot2) 
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(dev.args=list(bg="transparent"))
```

# Data Basics

### Data Basics

- \alert{Data} are information with context
- Data values, or \alert{observations} describe information about some entity
- \alert{Metadata} describe the process about how data is collected 

### Data Basics II

- \alert{Individuals} are the entities described by a set of data 
    - e.g. persons, households, firms, countries 
- \alert{Variables} are particular characteristics about an individual
    - e.g. age, income, profits, population, GDP, marital status, type of legal institutions
- \alert{Observations} are the individuals described by a collection of variables 
    - e.g. for one individual, we have their age, sex, income, education, etc.  
    - individuals and observations are *not necessarily* the same: 
        - e.g. we can have separate observations on the same individual over time 

\begin{figure}
		\includegraphics[height=1.35in]{individual1.jpg}
		\includegraphics[height=1.35in]{bargraph}
\end{figure}
        
### Categorical Variables 

- \alert{Categorial variables} place an individual into one of several possible categories
    - e.g. sex, season, political party
    - may be responses to questions
    - can be quantitative (e.g. age, zip code)

\begin{figure}
		\includegraphics[height=1.5in]{categoricaldata}
\end{figure}

### Representing Categorical Variables

\begin{table}
  \centering 
  \begin{tabular}{lrrrrr}
  Cut & Fair & Good & Very Good & Premium & Ideal\\ \toprule
  Count & 1610 & 4906 & 12082 & 13791 & 21551 \\ \midrule
  Proportion & 0.030 & 0.091 & 0.224 & 0.256 & 0.400\\ \bottomrule 
  \end{tabular}
  \caption*{Cut characteristics of 53,940 diamonds}
\end{table}
- A good way to represent categorical variables is with a \alert{frequency table} 
- **Count**: frequency (total number) of individuals in a category
- **Proportion**: *relative* frequency (percentage of all individuals) in a category

### Representing Categorical Variables II

```{r, diamonds, echo=TRUE, fig.height=3}
ggplot(diamonds, aes(x=cut))+
  geom_bar(fill="indianred", color="black")
```

- Charts and graphs are *always* better ways to visualize data
- A \alert{bar chart} represents categories as bars, with lengths proportional to the count or relative frequency fo each category

### Quantitative Variables

- \alert{Quantitative variables} take on numerical values of equal units that describe an individual
    - Units: points, dollars, inches 
    - Context: GPA, prices, height
- We mathematically manipulate quantitative variables *only* (even if categorical variables are numbers!)
    - e.g. sum, average, standard deviation

\begin{figure}
		\includegraphics[height=1.5in]{mathoperations}	
\end{figure}

### Context! 

- How variables are classified depends on the *purpose* of collecting and using the data

\onslide<2-> \begin{example}
  \begin{itemize}
    \item \emph{Age}, measured in years (quantitative) vs. categories of child, adult, senior, etc. 
  \end{itemize}
\end{example}

### Categorical or Quantitative?

\begin{example}
	What kind of data (categorical or quantitative) does each variable describe ?
	\begin{enumerate}[<+->]
		\item The number of pairs of shoes you own
		\item The type of car you drive
		\item Where you go on vacation
		\item The amount of money spent on a Super Bowl ad
		\item Customer ratings 
		\item The date a purchase was made
		\item Transaction ID	
		\item Number of correct answers on an exam
	\end{enumerate}
\end{example}

### Discrete Data

- \alert{Discrete data} are finite, with a countable number of alternatives 
    - Categorical: e.g. letter grades A, B, C, D, F
    - Quantitative: integers, e.g. SAT Score, number of children 

\begin{figure}
  \includegraphics[height=2in]{buildingblocks}
\end{figure}

### Continuous Data

- \alert{Continuous data} are infinitely divisible, with an uncountable number of alternatives
    - e.g. weights, temperature, GPA
- Many discrete variables may be treated as if they are continuous
    - e.g. SAT scores, wages 

\begin{figure}
		\includegraphics[height=1.5in]{continuous}	
\end{figure}

### Discrete or Continous?

\begin{example}
	What kind of data (discrete or continuous) does each variable describe ?
	\begin{enumerate}[<+->]
		\item Weight in pounds
		\item Price in dollars 
		\item Grade (Letter) 
		\item Temperature
		\item Amazon Star Rating 
		\item Number of customers
		\item Transaction ID	
		\item Number of correct answers on an exam
	\end{enumerate}
\end{example}


### Spreadsheets

```{r, example.spreadsheet.setup, echo=FALSE}
ID<-c(1,2,3,4,5)
Name<-c("John","Emile","Natalya","Lakisha","Cheng")
Age<-c(23,18,28,31,36)
Sex<-c("Male","Male","Female","Female","Male")
Income<-c(41000,52600,48000,60200,81900)
example<-data.frame(ID,Name,Age,Sex,Income)
```

```{r, example.spreadsheet, echo=FALSE}
example 
```

- The most common data structure we use is a \alert{spreadsheet}
    - Note: *R* calls this a `data frame`
- A **row** contains data about all variables for a single individual
- A **column** contains data about a single variable across all individuals
- It is good practice to have an **ID variable** to count and keep track of each observation

### Some Notation

- It is common to some notation like the following:
- Let $\{x_1, x_2, \cdots, x_n\}$ be a simple data series
    - $n$ individual observations
    - $x_i$ is the value of the $i$^th^ observation for $i=1,2,\cdots, n$
    
\begin{example}
 \begin{itemize}[<+->]
    \item Let $x$ represent the score on a homework assignment:
    \begin{equation*}
    75, 100, 92, 87, 79, 0, 95
    \end{equation*}
    \item What is $n$? 
    \item What is $x_1$?
    \item What is $x_6$?
  \end{itemize} 
\end{example}

### Datasets: Cross-Section

```{r, example.spreadsheet2, echo=FALSE}
example 
```

- \alert{Cross-sectional data}: observations of individuals at a given point in time
    - Each observation is a unique individual
    - Simplest and most common data 
    - A "snapshot" to compare differences across individuals
    
### Datasets: Time Series

```{r, example.timeseries.setup, echo=FALSE}
Year<-c(1950,1960,1970,1980,1985)
GDP<-c(8.2,9.9,10.2,12.4,13.6)
Unemployment<-c(0.06,0.04,0.08,0.08,0.06)
CPI<-c(100,118,130,190,196)
example.timeseries<-data.frame(Year,GDP,Unemployment,CPI)
```

```{r, example.timeseries, echo=FALSE}
example.timeseries
```

- \alert{Time-series data}: observations of the same individuals over time 
    - Each observation is an individual-year
    - Often used for macroeconomics, finance, and forecasting
    - Unique challenges for time series 
    - A "moving picture" to see how individuals change over time 

### Datasets: Panel Data

```{r, example.panel.setup, echo=FALSE}
City<-c("Philadelphia","Philadelphia","Washington D.C.","Washington D.C.", "New York")
Year<-c(1986, 1990, 1986, 1990, 1986)
Murders<-c(5, 8, 2, 10, 3)
Population<-c(3.7,4.2,0.250,0.275,6.4)
Unemployment<-c(8.7,7.2,5.4,5.5,9.6)
Police<-c(440,471,75,85,102)
example.panel<-data.frame(City,Year,Murders,Population,Unemployment,Police)
```

```{r, example.panel, echo=FALSE}
example.panel
```

- \alert{Panel dataset}, or \alert{longitudinal dataset}: a time-series for *each* cross-sectional entity
    - Must be the *same* cross-sectional entities over time 
    - More common today for serious researchers 
    - Unique challenges for panel data
    - A combination of "snapshot" comparisons and differences over time 

# Basic Statistics
 
### Variables and Distributions

- Variable shave a \alert{distribution} of different individual values (and how frequently they take these values)
- We want to *visualize* and *analyze* distributions to search for meaningful patterns using **statistics** 

### Two Types of Statistics 

- Two main categories or uses of statistics:
    1. \alert{Descriptive Statistics:} describes or summarizes the properties of a sample
		2. \alert{Inferential Statistics:} uses a sample in order to infer properties about a larger population

\begin{figure}
		\includegraphics[height=1.5in]{statsgraphs}	
\end{figure}

### Histograms 

- A common way to present a variable's distribution is a \alert{histogram}
    - The quantitative analog to the bar graph for a categorical variable 
- We divide up the data values into **bins** of a certain size, and count the number of values falling within each bin, representing them visually as bars 

### Histograms: Example

\begin{example}
A class of 13 students takes a quiz (out of 100 points) with the following results:

\begin{equation*}
\{ 0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95 \}
\end{equation*}
\end{example}

### Histogram: Example

```{r histogram2, echo=FALSE, fig.height=4.5, fig.width=10, warning=FALSE}
quizzes<-data.frame(scores=c(0,62,66,71,71,74,76,79,83,86,88,93,95))

quizzes.histogram<-ggplot(quizzes,aes(x=scores))+
  geom_histogram(breaks=c(0,10,20,30,40,50,60,70,80,90,100),color="black",fill="#56B4E9")+scale_x_continuous(breaks=c(0,10,20,30,40,50,60,70,80,90,100))+
  xlab("Scores")+ylab("Number of Students")
quizzes.histogram 
```

### Descriptive Statistics

- We are often interested in the *shape* or *pattern* of a distribution, particularly: 
    - Measures of central tendency
    - Measures of dispersion
    - Shape of distribution 

### Central Tendency: Mode

- The \alert{mode} of a variable is simply its most frequent value 
- A variable can have multiple modes 

\begin{example}
\{0, 62, 66, \textcolor<4->{magenta}{71}, \textcolor<4->{magenta}{71}, 74, 76, 79, 83, 86, 88, 93, 95\} 
\end{example}


### Multi-Modal Distributions

```{r, bimodal.example, echo=FALSE, fig.height=4.5}
quizzes.bimodal<-data.frame(scores=c(0,33,33,33,33,35,62,66,71,71,74,76,79,83,86,88,93,95))

quizzes.histogram.bimodal<-ggplot(quizzes.bimodal,aes(x=scores))+
  geom_histogram(breaks=c(0,10,20,30,40,50,60,70,80,90,100),color="black",fill="#56B4E9")+scale_x_continuous(breaks=c(0,10,20,30,40,50,60,70,80,90,100))+
  xlab("Scores")+ylab("Number of Students")
quizzes.histogram.bimodal

```
- Looking at a histogram, the modes are often the "peaks" of the distribution
- May be **unimodal**, **bimodal**, **trimodal**, etc

### Symmetry

```{r, symmetric.histogram, echo=FALSE, fig.height=4.5}

symmetric<-data.frame(x=rnorm(500,25,10))

symmetric.histogram<-ggplot(symmetric,(aes(x)))+
    geom_histogram(stat="bin",bins=10,color="black",fill="green")

symmetric.histogram
```

- A distribution is **symmetric** if it looks roughly the same on either side of the "center"
- The thinner ends (far left and far right) are called the **tails** of a distribution

### Symmetry and Skew 

```{r, skew.left.histogram, echo=FALSE, fig.height=4.5}

skew.left<-data.frame(x=rbeta(500,5,1))

skew.left.histogram<-ggplot(skew.left,(aes(x)))+
    geom_histogram(stat="bin",bins=10,color="black",fill="indianred")

#hist(rbeta(10000,5,2))
skew.left.histogram
```
- If one tail stretches farther than the other, distribution is **skewed** in the direction of the longer tail

### Outliers

- An extreme value that does not appear part of the general pattern of a distribution is an \alert{outlier}
```{r outlier, echo=FALSE, fig.height=3.5}
quizzes.histogram 
```

### Outliers

- Outliers can strongly affect descriptive statistics about a dataset
- Outliers can be the most informative part of the data
- Outliers could be the result of errors 
- Outliers should always be discussed in presentations about data

## Measures of Central Tendency

### Arithmetic Mean (Population)

- The natural measure of the center of a *population*'s distribution is its "average" or \alert{arithmetic mean ($\mu$)} 

$$\mu=\frac{x_1+x_2+...+x_N}{N} = \frac{1}{N} \sum^N_{i=1} x_i$$ 

- For $N$ values of variable $x$, "mu" is the sum of all individual $x$ values $(x_i)$ from 1 to $N$, divided by the $N$ number of values

### Arithmetic Mean (Sample)

- When we have a *sample*, we compute the \alert{sample mean ($\bar{x}$)} 

$$\bar{x}=\frac{x_1+x_2+...+x_n}{n} = \frac{1}{n} \sum^n_{i=1} x_i$$
-For $n$ values of variable $x$, "$x$-bar" is the sum of all individual $x$ values $(x_i)$ from 1 to $n$, divided by the $n$ number of values 

### Sample Mean Example
$$\{0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}$$

\begin{align*}
\bar{x}&=\frac{1}{13}(0+62+66+71+71+74+76+79+83+86+88+93+95)\\
&=\frac{944}{13}\\
&=72.61\\
\end{align*}

- Note the mean need not be an actual value of the data!

### Sample Mean and Outliers

$$\{62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}$$

- If we drop the outlier (0)

\begin{align*}
\bar{x}&=\frac{1}{12}(62+66+71+71+74+76+79+83+86+88+93+95)
&=\frac{944}{12}\\
&=78.67\\
\end{align*}

### Median
$$\{0, 62, 66, 71, 71, 74, \textcolor<4->{magenta}{76}, 79, 83, 86, 88, 93, 95\}$$

- The \alert{median} is the midpoint of the distribution
- 50% to the left of the median, 50% to the right of the median
- Arrange values in numerical order
    - For odd $n$: median is middle observation
    - For even $n$: median is average of two middle observations

### Median and Outliers
$$\{0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}$$
- The median is *robust* to outliers (if 0 changes to 62)
$$\{62, 62, 66, 71, 71, 74, \textcolor{magenta}{76}, 79, 83, 86, 88, 93, 95\}$$

### Mean, Median, Symmetry, Skew

$$\{1, 2, 2, 3, 3, 3, 4, \textcolor<3->{magenta}{4, 4}, 4, 5, 5, 5, 6, 6, 7\}$$ 

```{r, symmetric1, echo=FALSE, fig.height=3}
symmetric1<-data.frame(x=c(1,2, 2, 3, 3, 3, 4, 4, 4, 4, 5, 5, 5, 6, 6, 7))

symmetric1.histogram<-ggplot(symmetric1,aes(x=x))+
  geom_histogram(breaks=c(0,1,2,3,4,5,6,7),color="black",fill="#56B4E9")+scale_x_continuous(breaks=c(0,1,2,3,4,5,6,7))
symmetric1.histogram

```

- Mean: $\frac{64}{16}=4$
- Median: 4
- For a symmetric distribution, mean=median

### Mean, Median, Symmetry, Skew II

$$\{1, 2, 3, 4, 4, 4, \textcolor<3->{magenta}{5}, 5, 6, 6, 6, 7, 7\}$$ 

```{r, skewl, echo=FALSE, fig.height=3}
leftskew<-data.frame(x=c(1,2,3,4,4,4,5,5,6,6,6,7,7))

leftskew.hist<-ggplot(leftskew,aes(x=x))+
  geom_histogram(breaks=c(0,1,2,3,4,5,6,7),color="black",fill="#56B4E9")+scale_x_continuous(breaks=c(0,1,2,3,4,5,6,7))
leftskew.hist
```

- Mean: $\frac{60}{13}=4.6$
- Median: 5
- For a left-skewed distribution, mean$<$median

### Mean, Median, Symmetry, Skew III

$$\{1, 1, 2, 2, 2, 3, \textcolor<3->{magenta}{3}, 4, 4, 4, 5, 6, 7\}$$ 

```{r, skewr, echo=FALSE, fig.height=3}
rightskew<-data.frame(x=c(1,1,2,2,2,3,3,4,4,4,5,6,7))

rightskew.hist<-ggplot(rightskew,aes(x=x))+
  geom_histogram(breaks=c(0,1,2,3,4,5,6,7),color="black",fill="#56B4E9")+scale_x_continuous(breaks=c(0,1,2,3,4,5,6,7))
rightskew.hist
```

- Mean: $\frac{44}{13}=3.4$
- Median: 3
- For a right-skewed distribution, mean$>$median

## Measures of Spread

### Measures of Spread

- The more *variation* in the data, the less helpful a measure of central tendency will tell us
- Beyond just the center, we also want to measure the spread
- Simplest metric is \alert{range}=max-min

### Five Number Summary

Once we know the values of the quartiles, we can construct a \alert{five-number summary} of a distribution, including:
    1. Minimum
    2. $Q_1$ (25%)
    3. Median (50%)
    4. $Q_3$ (75%)
    5. Maximum
    
### Five Number Summary II

```{r fivenum}
summary(quizzes$scores)
```

### Boxplots

- Graphical way to visualize five number summary is a \alert{boxplot}
\begin{columns}	
	\begin{column}[c]{8cm}
		\begin{itemize}
				\item<3-> The length of the box is the IQR (Q1-Q3) 
				\item<4-> The line within the box is the median 
				 \item<5-> The ``whiskers'' identify data within $1.5 \times IQR$
				\item<6-> Points beyond the whiskers are \textcolor{magenta}{outliers} 
		\end{itemize}
		\end{column}
		\begin{column}[c]{4cm}
			\begin{figure}
			\begin{tikzpicture}[scale=0.45]\scriptsize 
				\draw<3->[black, thick, fill=green!50] (0,0)--(0,2)node[left]{$Q_3$}--(2,2)--(2,0);
				\draw<3->[black, thick, fill=red!50] (0,0)--(0,-3)node[left]{$Q_1$}--(2,-3)--(2,0);
				\draw<4->[very thick, dashed, black] (0,0)--(2,0)node[right]{Median};
				\draw<5->[black, thick] (1,2)--(1,4)--(0.75,4)--(1.25,4)node[right]{$Q_3+1.5*IQR$};
				\draw<5->[black, thick] (1,-3)--(1,-4)--(0.75,-4)--(1.25,-4)node[right]{$Q_1-1.5*IQR$};
				\draw<6->[magenta, fill=magenta] (1,5) circle (0.125cm)node[right]{Outlier};
			\end{tikzpicture}
		\end{figure}
		\end{column}
\end{columns}

### Boxplots II

\begin{center} Quiz 1: \{0, 62, 66, 71, 71, 74, 76, 79, 83, 86, 88, 93, 95\}	\\
			Quiz 2: \{50, 62, 72, 73, 79, 81, 82, 82, 86, 90, 94, 98, 99\} \end{center} 
		\onslide<2->Quiz 1 
\begin{table}
			\begin{tabular}{ccccc}
			Min & $Q_1$ & Median & $Q_3$ & Max\\ \toprule 
			0 & 71 & 76 & 86 & 95\\ \bottomrule 	
			\end{tabular}
		\end{table}
		\onslide<3->Quiz 2
			\begin{table}
			\begin{tabular}{ccccc}
			Min & $Q_1$ & Median & $Q_3$ & Max\\ \toprule 
			50 & 73 & 82 & 90 & 99\\ \bottomrule 	
			\end{tabular}
\end{table}

### Boxplots III

```{r second.quiz, echo=FALSE, fig.height=4}
quizzes<-data.frame(scores=c(0,62,66,71,71,74,76,79,83,86,88,93,95))
#rename scores as quiz 1
colnames(quizzes)<- c("quiz1")
#add student variable to track student over multiple quizzes
quizzes$student<-(1:13)

#add quiz 2
quizzes$quiz2<-c(50, 62, 72, 73, 79, 81, 82, 82, 86, 90, 94, 98, 99)

# second quiz boxplot
library("tidyr") #for gather 
quizzesl<-gather(quizzes, key="student", value="scores") #need to create a variable for which quiz it is (x axis below)
colnames(quizzesl)<-c("quiz","score")
quizzes.boxplot<-ggplot(quizzesl, aes(x=quiz,y=score,fill=quiz))+
    geom_boxplot()+
    stat_boxplot(geom='errorbar')+
    xlab("Quiz")+ylab("Scores")
quizzes.boxplot
```


### Deviations

- Each observation \textcolor{magenta}{deviates} from the mean of the data: 
$$deviation = x_i-\mu	$$

- There are as many deviations as there are data points ($n$)
- We can measure the \emph{average} or \textcolor{magenta}{standard deviation} from the mean

### Variance 

- The \textcolor{teal}{population variance ($\sigma^2$)} of a \emph{population} distribution measures the average of the \emph{squared} deviations from the population mean

$$\sigma^2 = \frac{1}{N}\displaystyle\sum^N_{i=1} (x_i-\mu)^2	$$

- Why do we square deviations? 

### Standard Deviation

-Square root the variance to get the \textcolor{teal}{population standard deviation ($\sigma$)}, the average deviation from the mean (in $x$ units)
$$\sigma=\sqrt{\sigma^2}=\sqrt{\frac{1}{N}\displaystyle\sum^N_{i=1} (x_i-\mu)^2	}$$

### Sample Variance

- The \textcolor{magenta}{sample variance ($s^2$)} of a \emph{sample} distribution measures the average of the \emph{squared} deviations from the sample mean

$$s^2 = \frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2	$$


-Why divide by $n-1$?

### Sample Standard Deviation
- Square root the variance to get the \textcolor{teal}{sample standard deviation ($s$)}, the average deviation from the mean (in $x$ units)

$$s=\sqrt{s^2}=\sqrt{\frac{1}{n-1}\displaystyle\sum^n_{i=1} (x_i-\bar{x})^2	}$$

### Sample Standard Deviation: Example

\begin{example}
Calculate the sample standard deviation for the following series: 
$$\{2, 4, 6, 8, 10 \}$$
\end{example}

### Descriptive Statistics: Population vs. Sample

\begin{columns}
		\begin{column}[c]{5cm}
		Population Parameters
		\begin{itemize}
			\item Population Size: $N$
			\item Mean: $\mu$
			\item Variance: $\sigma^2=\frac{1}{N} \displaystyle\sum^N_{i=1} (x_i-\mu)^2$
			\item Standard Deviation: $\sigma = \sqrt{\sigma^2}$
		\end{itemize}
		\end{column}
		\begin{column}[c]{5cm}
		Sample Statistics
		\begin{itemize}
			\item Sample Size: $n$
			\item Mean: $\bar{x}$
			\item Variance: $s^2=\frac{1}{n-1} \displaystyle\sum^n_{i=1} (x_i-\bar{x})^2$
			\item Standard Deviation: $s = \sqrt{s^2}$
		\end{itemize}
		\end{column}
\end{columns}


