---
title: "Lecture 9: Regression Diagnostics and Plotting with `ggplot2`"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "September 26, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(ggplot2) 
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(dev.args=list(bg="transparent"))
```

```{r, echo=FALSE}
# load data set and run regression again 
library("foreign") #for importing .dta files
CASchool<-read.dta("~/Dropbox/Teaching/Hood College/ECON 480 - Econometrics/Data/caschool.dta")
school.regression<-lm(testscr~str, data=CASchool)
```

# Solvable Regression Problem \#1: Heteroskedasticity

### Homoskedasticity 

- Recall assumption \#2 about the regression residuals ($\epsilon$) are that they are \alert{homoskedastic}:

$$var(\epsilon)=\sigma_{\epsilon}^2$$

- A fancy way of saying that the **variance of the residuals is constant**, i.e. does not change over values of $X$
- Combined with assumption \#1 (the mean of the residuals $E[\epsilon]=0$) $\implies$ residuals are i.i.d. and come from the same distribution $\sim(0,\sigma^2_{\epsilon}$)

### Homoskedasticity

- If residuals are \alert{heteroskedastic}, they do *not* have the same variance over values of $X$

$$var(\epsilon) \neq \sigma_{\epsilon}^2$$

- **This does not cause $\hat{\beta_1}$ to be biased**, but it does cause the standard error of $\hat{\beta_1}$ to be incorrect
    - Recall, $se(\hat{\beta_1})$ is used to calculate our test statistic for hypothesis testing
    - **May overstate the statistical significance of a finding!**

### Homoskedasticity

- The formula for $se(\hat{\beta_1})$ assumes homoskedasticity, recall (from Lecture 8) it was:

$$se(\hat{\beta_1})=\sqrt{\frac{(SER)^2}{n \times var(X)}}$$

- When errors are heteroskedastic, the formula mutates to

\onslide<2-> 
$$se(\hat{\beta_1})=\sqrt{\frac{\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2\hat{\epsilon}^2}{\big[\displaystyle\sum^n_{i=1}(X_i-\bar{X})^2\big]^2}}$$

- This is \alert{heteroskedasticity-robust} ("\alert{robust}") method of calculating $se(\hat{\beta_1})$
- Don't learn formula, **do learn what heteroskedasticity is and how it affects our model!**

### Visualizing Heteroskedasticity


```{r, echo=FALSE, fig.height=4}
# Let's make some heteroskedastic data

#runif generates random numbers via a random uniform distribution (# of draws, lower lim, upper lim) 
x<-runif(500,0,10)

#rnorm generates random numbers via a normal distribution (# of draws, mean, sd)
# we will include x as the lower and upper limit 
y<-rnorm(500,x,x)

het.data<-data.frame(x,y)

# plot data 
library(ggplot2)
het.plot<-ggplot(het.data, aes(x,y))+
  geom_point(color="blue")+
  geom_smooth(method=lm, color="red")+
  scale_x_continuous(breaks=c(0,2,4,6,8,10))+
  theme_bw()

het.plot

```

- The average residual (distance from point to OLS line) changes as $X$ changes

### Visualizing Heteroskedasticity II 

```{r,echo=FALSE,fig.height=4}
het.plot+geom_vline(xintercept=c(0,2,4,6,8,10),linetype="dashed",color="black",size=1)
```

- We would expect the distribution of the residuals ($\hat{\epsilon}$) to be the same at every value of $X$
- Clearly there are very different distributions of residuals across $X$

### Visualizing Heteroskedasticity III 

```{r,echo=FALSE,fig.height=4}
het.reg<-lm(y~x,het.data) # run regression

het.data$res<-residuals(het.reg) # generate residuals, store in het.data

# define a variable called range in the het.data dataframe

# make 5 groups, (called 2,4,6,8,10) for range 
  # values are in group 2 if x is between 0 and 2
  # etc for groups 4, 6, 8, 10
het.data$range[het.data$x>0 & het.data$x<2] = 2
het.data$range[het.data$x>2 & het.data$x<4] = 4
het.data$range[het.data$x>4 & het.data$x<6] = 6
het.data$range[het.data$x>6 & het.data$x<8] = 8
het.data$range[het.data$x>8 & het.data$x<10] = 10

# we want to make range a factor (categorical) variable
# to plot it below as a series of groups

het.data$range<-as.factor(het.data$range)

# plot density by range group

ggplot(het.data, aes(x=res,fill=range))+
  geom_density()+facet_grid(cols=vars(het.data$range))+
  xlab("Residuals by Range of X")+ylab("Probability Density")+
  scale_fill_discrete(name="Residuals for",
                         labels=c("0<X<2", "2<X<4", "4<X<6",
                                  "6<X<8", "8<X<10"))

```

- We would expect the distribution of the residuals ($\hat{\epsilon}$) to be the same at every value of $X$
- Clearly there are very different distributions of residuals across $X$

### Visualizing Heteroskedasticity IV 

```{r, echo=FALSE, fig.height=4}
ggplot(het.data,aes(x=x,y=res))+
  geom_point(color="red")+geom_hline(yintercept=0, color="blue")+
  xlab("X")+ylab("Residuals")
```

- We can also see in the \alert{residual plot} that the size of residuals increases as $X$ increases

### Detecting Heteroskedasticity

- Several tests to check if data is heteroskedastic 
- One common test is **Breusch-Pagan test** 
- Can use `bptest()` with `lmtest` package in `R`
    - $H_0$: homoskedastic
    - If $p$-value < 0.05, reject $H_0\implies$ heteroskedastic

\onslide<6->

```{r, message=FALSE}
library("lmtest") #load lmtest package, install if first time
bptest(het.reg)
```

### Standard Errors: Technical Digression

- Before we generate \alert{robust SE's}, let's look at plain SEs
- Regression creates a **variance-covariance** matrix ('`vcov`') of OLS estimators (betas), e.g. 

\onslide<2->

$$\begin{bmatrix}
cov(\hat{\beta_0},\hat{\beta_0}) & cov(\hat{\beta_0},\hat{\beta_1}) \\
cov(\hat{\beta_0}, \hat{\beta_1}) & cov(\hat{\beta_1},\hat{\beta_1}) \\
\end{bmatrix}$$

- Why a matrix?
    - We currently have 2 OLS estimators ($\hat{\beta_0}, \hat{\beta_1}$) for 1 $X_i$ variable 
    - We eventually will have $k$ $X_i$ variables, so a $\big(k+1 \times k+1\big)$ matrix 
- The 'diagonal' of the matrix contains the variance of each OLS estimator, since $cov(X,X)=var(X)$
- Taking the diagonal and square rooting the terms gives us the SE of each estimator

### Standard Errors: In `R` 

\scriptsize 

```{r}
# the variance-covariance matrix
vcov(het.reg)
#the var(beta)'s are the diagonal of the matrix

diag(vcov(het.reg)) # look at just the diagonal values

#convert into standard errors by square rooting 
sqrt(diag(vcov(het.reg)))
```

### Standard Errors: In `R` II 

\scriptsize 

```{r}
# confirming the SE's match what R finds automatically with lm()
summary(het.reg)
```

### Using Robust Standard Errors in `R` 

\footnotesize 

```{r, warning=FALSE, message=FALSE}
library("sandwich") # package that allows for robust SE estimation, install if first use
library("lmtest") # package that allows for coefficient tests, install if first use

# take original regression and change standard errors to robust SEs # 

# create Robust Standard Errors for regression as 'het.reg$rse' 
het.reg$rse <-sqrt(diag(vcovHC(het.reg, type="HC1"))) 
# same procedure as above but now we generate vcov with "HC1" method (technical)

```

### Using Robust Standard Errors in `R` II

- Using `coeftest()` function in the `lmtest` package, hypothesis tests with robust SEs

\tiny  

```{r}
coeftest(het.reg) # test with normal SEs 
coeftest(het.reg,vcov=vcovHC(het.reg,"HC1")) # tests with robust SEs
```


### Using Robust Standard Errors in `R`: `stargazer` output 

\center 

```{r,echo=FALSE,results="asis", message=FALSE, warning=FALSE}

library("stargazer")
stargazer(het.reg, het.reg, se=list(NULL,het.reg$rse), type="latex", header=FALSE, style="qje", 
          column.labels = c("Normal SEs", "Robust SEs"), title="Regression Results", float=FALSE, 
          font.size = "scriptsize", omit.stat=c("adj.rsq","f")) 
```


# Solvable Regression Problem \#2: Outliers

### Outliers Can Bias OLS

```{r, echo=FALSE, fig.height=4}
#Just to simplify the dataset, let's make a new, smaller data frame with just the variables: observation, district, testscr, and str 
CA.simple<-subset(CASchool,select=c(observat,district,testscr,str))

#Now let's create two crazy outliers 
outlier<-c(421,"Terrible School 1",800,30)
outlier2<-c(422,"Terrible School 2",850,28)
outlier3<-c(423,"Terrible School 2",820,29)

#And add them to a new dataframe 
CA.outlier<-rbind(CA.simple,outlier,outlier2,outlier3)
CA.outlier$str<-as.numeric(CA.outlier$str) #dont know why it changes class to character, change to numeric
CA.outlier$testscr<-as.numeric(CA.outlier$testscr)
outliers<-subset(CA.outlier, testscr>790) 

#Create a scatterplot with the outlier 
ggplot(CA.simple, aes(str,testscr))+
  geom_point(color="blue")+ #the normal data points
  geom_smooth(data=CA.simple,method=lm, color="red")+ #the regression line
  geom_point(data=outliers, color="magenta")+ #the outliers (in magenta)
  geom_smooth(data=CA.outlier, method=lm,color="purple")+ #the regression line with outliers
  xlab("Student to Teacher Ratio")+ylab("Test Score")+theme_bw()
```

- \alert{Outliers} can affect the slope (and intercept) of the line

### Outliers Can Bias OLS

\center 

```{r, echo=FALSE, warning=FALSE, message=FALSE, results="asis"}
# run regression with outlier
school.outlier.reg<-lm(testscr~str, data=CA.outlier)

# output plain regression and outlier regression 
stargazer(school.outlier.reg, school.regression, type="latex", header=FALSE, style="qje", 
          column.labels = c("With Outliers", "Without Outliers"), title="Regression Results", float=FALSE, 
          font.size = "scriptsize", omit.stat=c("adj.rsq","f")) 
```


### Detecting Outliers

- Plot the data and look! 
- A few methods to detect \alert{influence}: ability of individual observations to affect OLS estimates

### Detecting Outliers with `car` Package

```{r, message=FALSE, warning=FALSE}
library("car")

# Use Bonferonni test 
outlierTest(school.outlier.reg) # will point out which obs #s seem outliers
```

### dfbetas

- **dfbetas** measure how different each OLS coefficient will be without each observation
    - Scales the measure by standard error of OLS coefficient with observation deleted
    - Name means "difference in beta" from deleted observation
        - e.g. $dfbeta_i=-3$: observation $i$ decreases coefficient by 3 standard errors 
    - Downside: calculates this measure for *each* observation for *each* beta ($n \times k$ dfbetas)! 

### dfbetas in R

\tiny 
```{r}
dfbetas(school.outlier.reg)
```

### Dealing with Outliers

- Often, outliers may be the result of human error (measurement, transcribing, etc)
- Outliers may be meaningful and accurate
- In any case, compare how including/dropping outliers affects regression and always discuss outliers! 

# Advanced Plotting in `R` with `ggplot2`

### ``ggplot2``

\begin{columns}
  \begin{column}[c]{9cm}
    \begin{itemize}[<+->]
      \item \texttt{ggplot2} is the most popular package
      \item Created by Hadley Wickham, part of \texttt{tidyverse} (really the "Hadleyverse")
      \item Very powerful and beautiful, but requires a bit more of a learning curve
      \item All those "cool graphics" you've seen in the New York Times, fivethirtyeight, the Economist, Vox, etc. used ggplot2 
      \item \texttt{gg} stands for a \textbf{grammar of graphics}
    \end{itemize}
  \end{column}
  \begin{column}[c]{4cm}
    \begin{figure}
      \includegraphics[height=2.5in]{hadleyplot2}
    \end{figure}
  \end{column}
\end{columns}

### A Picture is Worth a Thousand Words

\begin{figure}
  \includegraphics[height=2.25in]{racialdotnyc}
  \caption*{\footnotesize Dustin Cable's Racial Dot Map of NYC\footnote{1 dot = 1 person, colors: \textcolor{blue}{White}, \textcolor{green}{African-American}, \textcolor{red}{Asian}, \textcolor{orange}{Latino}, \textcolor{brown}{All Other}}, \href{https://www.wired.com/2013/08/how-segregated-is-your-city-this-eye-opening-map-shows-you/}{The Best Map Ever Made of America's Racial Segretation}; his (Python) code is open-source and available on \href{https://github.com/unorthodox123/RacialDotMap}{Github}}
\end{figure}

### `ggplot2`: All Your Figure Are Belong to Us

\begin{table}
  \begin{tabular}{ccc}
    \includegraphics[height=1.4in]{538gg1} & \includegraphics[height=1.4in]{538gg2} & \includegraphics[height=1.4in]{nytgg}\\
    From \href{https://fivethirtyeight.com/features/fastest-airlines-fastest-airports/}{fivethirtyyeight} & From \href{https://fivethirtyeight.com/features/fastest-airlines-fastest-airports/}{fivethirtyyeight} & From \href{https://archive.nytimes.com/www.nytimes.com/interactive/sports/football/2013-fantasy-football-tier-charts-QB.html}{New York Times} \\
  \end{tabular}
\end{table}

### `ggplot2` figures 

\begin{columns}
  \begin{column}[c]{8cm}
    \begin{itemize}
      \item Figures produced are \textbf{vector-based} graphics (.pdf, .svg)
      \begin{itemize}
        \item Can rescale to any size and not look "pixely"
        \item But big file size
      \end{itemize}
      \item<2-> Can save figures as \textbf{raster-based} graphics (.png, .jpeg, .bmp, .gif)
      \begin{itemize}
        \item Look \emph{awful} when blown up
        \item But small file size, great for MS Office and Web
      \end{itemize}
      \item<3-> Suggestions:
      \begin{itemize}
        \item If printing on paper, save graphics as .pdf 
        \item If posting to the web, save as .png and specify size 
      \end{itemize}
    \end{itemize}
  \end{column}
  \begin{column}[c]{4cm}
    \begin{figure}
      \includegraphics[height=2in]{vectorgraphics}
    \end{figure}
  \end{column}
\end{columns}

### Two Ways to Plot

1. Just the single `ggplot` command 
    - Will view plot right after producing it
    - Does not save as an object
    - Need to rerun or copy/paste full command producing plot in order to modify or view it again
    - Can still put it in a document


```{r, eval=FALSE}
ggplot(...) # make and view plot
ggplot(some.options) # remake plot with new options and view plot
```

### Two Ways to Plot II 

2. Create an object (as usual in `R`)
    - This allows you to save the plot for later (re)use
    - Also allows you to modify it 
    - Any time you want to view display it (i.e. for putting it in a document), just call up the plot by name

```{r, eval=FALSE}
plot.name<-ggplot(...) # make plot
plot.name<-plot.name+some.options # add new options to existing plot
plot.name # view plot 
```

### Grammar of Graphics

```{r, eval=FALSE}
plot.name<-ggplot(data=mydf, mapping=aes(x=xvar,y=yvar))+
  geom_something(options)+
  moreoptions...
```

- `gg` "grammar of graphics" implies any graphic can be built from the same components/layers:
    1. **Data**: base-layer describes the data used
        - `mydf` is the dataframe containing `xvar` and `yvar`
        - `aes()` "aesthetics" identifies `xvar` (and if applicable `yvar`) from data to be "mapped" to a visual mark
    2. **Geoms**: visual marks that represent data observations or models, common examples:
        - e.g. `geom_point`, `geom_line`, `geom_bar`, `geom_histogram`, `geom_density`, `geom_boxplot`
    3. **Coordinates**: Cartesian coordinates are default
        - change scales, axes, labels, etc; advanced options like maps 

### Grammar of Graphics II

- Most important idea to master is `aes()` **aesthetics** that map data to visual markings
- Aesthetics come in many forms and many options, depending on the context of the data
    - Must identify position (e.g. what is `x` and `y`) 
    - Determine the marking with various `geoms` (points, bars, lines, boxes, etc)
    - Can pass additional options into `geom` (color, size, shape, etc) 
        - Particularly important if we want color, size, or shape to depend on a particular variable in dataset

\begin{figure}
  \includegraphics[height=1.5in]{ggplotaes}
\end{figure}

### Example

For our example, we'll use the [mpg dataset](https://ggplot2.tidyverse.org/reference/mpg.html) loaded with the `ggplot2` package 

```{r}
library("ggplot2") #load ggplot2
mpg #look at dataset
```

### `gg` Histogram: Base Layer

- Start with the base layer: establish the data source, define $x$ variable

```{r, fig.height=3.5}
mpg.h<-ggplot(data=mpg,mapping=aes(x=hwy))
mpg.h
```

### `gg` Histogram: Adding Geoms

- Add a histogram layer of `hwy`

```{r, fig.height=3.5, message=FALSE, warning=FALSE}
mpg.h1<-mpg.h+geom_histogram()
mpg.h1
```

### `gg` Histogram: Customizing Geoms

- Edit the histogram (\# of bins, color, etc)

```{r, fig.height=3.5}
mpg.h2<-mpg.h+geom_histogram(bins=20, color="black",fill="indianred")
mpg.h2
```

### `gg` Histogram: Adding Other Layers

- Add a vertical line for the mean with another `geom` called `vline`

```{r, fig.height=3.5}
mpg.h2<-mpg.h2+
  geom_vline(xintercept=mean(mpg$hwy),linetype="dotted",color="blue",size=1)
mpg.h2
```

### `gg` Histogram: Editing Coordinates (Axes)

- Change the labels on the axes with `xlab()` and `ylab()`

```{r, fig.height=3.5}
mpg.h2<-mpg.h2+xlab("Miles Per Gallon (on Highway)")+ylab("Number of Cars")
mpg.h2
```

### `gg` Other Geoms

- How about a **density plot**: use `geom_density()` instead of `geom_histogram()`

```{r, fig.height=3.5}
mpg.d<-ggplot(data=mpg,aes(x=hwy))+
  geom_density(fill="indianred")
mpg.d
```

### `gg` Other Geoms

- Let's make a separate density plot for each `class`, set `aes` to `fill` by `class`

```{r, fig.height=3.5}
mpg.d<-ggplot(data=mpg,aes(x=hwy,fill=class))+
  geom_density(alpha=0.5) # alpha adds transparency
mpg.d
```

### `gg` Boxplot

- Instead of a density plot, a `boxplot` by `class` (note now `x` is `class` and `y` is `hwy`): 

```{r, fig.height=3.5}
mpg.b<-ggplot(data=mpg,aes(x=class,y=hwy,fill=class))+
  geom_boxplot()
mpg.b
```


### Scatterplot

- Start with the base layer: establish data source, define $x$ and $y$ variables

```{r, fig.height=3.5}
mpg.p<-ggplot(data=mpg,aes(x=displ, y=hwy)) #use mtcars df, let x=displ, y=hwy

mpg.p
```

### Scatterplot: Geom Layer

```{r, fig.height=3.5}
mpg.p<-mpg.p+geom_point() # specify observations as points on graph

mpg.p
```

### Scatterplot: Geom Layer Options

```{r, fig.height=3.5}
mpg.p<-mpg.p+geom_point(aes(color=manufacturer)) # color data points by manuf.

mpg.p
```

### Scatterplot: Geom Layer II 

```{r, fig.height=3.5}
mpg.p<-mpg.p+geom_smooth(method="lm", color="black") # add a black OLS line 

mpg.p
```

### Scatterplot: Coordinate Layer

```{r, fig.height=3.5}
mpg.p<-mpg.p+xlab("Engine Displacement (Liters)")+
  ylab("Miles Per Gallon on Highway")
mpg.p
```

### Scatterplot: Coordinate Options

- Let's have some fun changing the theme 

```{r, fig.height=3.5}
library("ggthemes") # need ggthemes package (install if first use)
mpg.p<-mpg.p+theme_economist_white() #make it look like The Economist magazine
mpg.p
```


### Scatterplot: Coordinate Options II 

```{r, fig.height=3.5}
mpg.p<-mpg.p+theme_fivethirtyeight() #make it look like fivethirtyeight
mpg.p
```

### Scatterplot: Coordinate Options: Facetting

```{r, fig.height=3.5}
# make columns of separate 'facet' figures for each class of car
mpg.p<-mpg.p+facet_grid(cols = vars(class)) # make 'columns' by variable 'class'
mpg.p
```

### All Together Now

```{r, fig.height=3.5}
ggplot(data=mpg,aes(x=displ, y=hwy))+geom_point(aes(color=manufacturer))+
  geom_smooth(color="black",method="lm")+
  xlab("Engine Displacement (Liters)")+ylab("Miles Per Gallon on Highway")+
  theme_fivethirtyeight()+facet_grid(cols = vars(class)) 
```

### Advanced Uses of `ggplot2`: Maps (See Rmd for Code)

```{r,echo=FALSE,fig.height=5.5,message=FALSE}
# requires packages "maps" and "reshape2", in addition to ggplot2
crimes <- data.frame(state = tolower(rownames(USArrests)),USArrests)
library(reshape2) # for melt
crimesm <- melt(crimes, id = 1)
# Get map data
require(maps) 
map_data <- map_data("state")
# Plot the map with Murder data
ggplot(crimes, aes(map_id = state)) + 
  geom_map(aes(fill = UrbanPop), map = map_data) + 
  expand_limits(x = map_data$long, y = map_data$lat)+
  xlab("Longitude")+ylab("Latitude")
```


### Advanced Uses of `ggplot2`: Maps (See Rmd for Code) II 

```{r, echo=FALSE, message=FALSE,warning=FALSE,fig.height=5.5}
dat <- read.csv("../Data/EconomistData.csv")
library("tidyverse")
library("ggrepel") # to push text labels away from actual point

mR2 <- summary(lm(HDI ~ CPI + log(CPI), data = dat))$r.squared
mR2 <- paste0(format(mR2, digits = 2), "%")

p <- ggplot(dat,
            mapping = aes(x = CPI, y = HDI)) +
  geom_smooth(mapping = aes(linetype = "r2"),
              method = "lm",
              formula = y ~ x + log(x), se = FALSE,
              color = "red") +
  geom_point(mapping = aes(color = Region),
             shape = 1,
             size = 4,
             stroke = 1.5) +
  geom_text_repel(mapping = aes(label = Country, alpha = labels),
                  color = "gray20",
                  data = transform(dat,
                                   labels = Country %in% c("Russia",
                                                           "Venezuela",
                                                           "Iraq",
                                                           "Mayanmar",
                                                           "Sudan",
                                                           "Afghanistan",
                                                           "Congo",
                                                           "Greece",
                                                           "Argentinia",
                                                           "Italy",
                                                           "Brazil",
                                                           "India",
                                                           "China",
                                                           "South Africa",
                                                           "Spain",
                                                           "Cape Verde",
                                                           "Bhutan",
                                                           "Rwanda",
                                                           "France",
                                                           "Botswana",
                                                           "France",
                                                           "US",
                                                           "Germany",
                                                           "Britain",
                                                           "Barbados",
                                                           "Japan",
                                                           "Norway",
                                                           "New Zealand",
                                                           "Sigapore"))) +
  scale_x_continuous(name = "Corruption Perception Index, 2011 (10=least corrupt)",
                     limits = c(1.0, 10.0),
                     breaks = 1:10) +
  scale_y_continuous(name = "Human Development Index, 2011 (1=best)",
                     limits = c(0.2, 1.0),
                     breaks = seq(0.2, 1.0, by = 0.1)) +
  scale_color_manual(name = "",
                     values = c("#24576D",
                                "#099DD7",
                                "#28AADC",
                                "#248E84",
                                "#F2583F",
                                "#96503F"),
                     guide = guide_legend(nrow = 1, order=1)) +
  scale_alpha_discrete(range = c(0, 1),
                       guide = FALSE) +
  scale_linetype(name = "",
                 breaks = "r2",
                 labels = list(bquote(R^2==.(mR2))),
                 guide = guide_legend(override.aes = list(linetype = 1, size = 2, color = "red"), order=2)) +
  ggtitle("Corruption and human development") +
  labs(caption="Sources: Transparency International; UN Human Development Report") +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = "gray"),
        text = element_text(color = "gray20"),
        axis.title.x = element_text(face="italic"),
        axis.title.y = element_text(face="italic"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.box = "horizontal",
        legend.text = element_text(size = 12),
        plot.caption = element_text(hjust=0),
        plot.title = element_text(size = 16, face = "bold"))
p

```

### Getting Help and Customizing to Your Heart's Content

\begin{columns}
  \begin{column}[c]{9cm}
    \begin{itemize}
      \item Every single command and option has \href{https://ggplot2.tidyverse.org/reference/index.html}{extensive documentation with examples}
      \begin{itemize}
        \item<2-> Can type \texttt{?} in front of any function to get help and examples (e.g. \texttt{?aes()}, \texttt{?geom\_smooth()})
        \item<3-> More help and examples from Wickham's \href{http://r4ds.had.co.nz/data-visualisation.html}{R for Data Science book chapter on ggplot2}
      \end{itemize}
      \item<4-> Data work is a science, but it should also be \textbf{art}!
      \item<5-> I will \emph{not}  grade you on the beauty of your graphics (but we're all secretly judging!)
      \begin{itemize}
        \item<6-> \texttt{hist()} and \texttt{plot()} are fine, you are not required to use \texttt{ggplot2} (but you really should! )
      \end{itemize}
    \end{itemize}
  \end{column}
  \begin{column}[c]{4cm}
    \begin{figure}
      \includegraphics[height=2.5in]{hadleyplot2}
    \end{figure}
  \end{column}
\end{columns}

### More Help

![[Rstudio's Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)](ggplot2cheat.png)


### Beautiful Graphics: Less is More 

\begin{figure}
  \includegraphics[height=2.5in]{badfigure}
  \caption*{"Kill me now"}
\end{figure}

### Beautiful Graphics: Less is More II 

\begin{figure}
  \includegraphics[height=2.5in]{goodfigure}
\end{figure}

### Beautiful Graphics: Less is More III  

\begin{figure}
  \includegraphics[height=2in]{goodfigure2}
\end{figure}

### Beautiful Graphics: Less is More IV

[Less is More: Example Gif](http://stat545.com/img/less-is-more-darkhorse-analytics.gif)

### Resources (Best are **Bolded**)

On `ggplot2`

- **R Studio's [ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)**
- **`ggplot2`'s website [reference section](https://ggplot2.tidyverse.org/reference/index.html)**
- STHDA's [be awesome in ggplot2](http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization)
- r-statistic's [top 50 ggplot2 visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
- Hadley Wickham's \href{http://r4ds.had.co.nz/data-visualisation.html}{R for Data Science book chapter on ggplot2}

On data visualization

- **Kieran Healy's [Data Visualization: A Practical Guide](http://socviz.co/)**
- **Claus Wilke's [Fundamentals of Data Visualization](https://serialmentor.com/dataviz)**
- PolicyViz [Better Presentations](https://policyviz.com/) 
- Karl Broman's [How to Display Data Badly](https://www.biostat.wisc.edu/~kbroman/presentations/IowaState2013/graphs_combined.pdf)
- [I Want Hue](http://tools.medialab.sciences-po.fr/iwanthue/) 
