---
title: "Lecture 16: Nonlinear and Polynomial Models"
subtitle: "ECON 480 - Econometrics - Fall 2018"
author: "Ryan Safner"
date: "November 26, 2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
#fontsize:bigger
---

```{r setup, include=FALSE}
# For making transparent ggplot2 graphs, from https://gist.github.com/cboettig/5600558 

# Set plotting to bw plot default, but with transparent background elements.  
# Note transparency requires the panel.background, plot.background, and device background all be set!
library(tidyverse) 
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, dev.args=list(bg="transparent"))
```

### Linear Regression

- OLS is commonly known as "**\emph{linear} regression**" as it fits a *straight line* to data points 
- Often, data and relationships between variables may *not* be linear 

### Nonlinearities? Example

```{r, echo=FALSE}
library("gapminder")
library("ggplot2")
```

```{r, fig.height=4, echo=FALSE}
p1<-ggplot(gapminder, aes(gdpPercap,lifeExp))+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_y_continuous(name="Life Expectancy (Years)", breaks=seq(0,90,20),limits=c(0,95))+
  scale_x_continuous(name="GDP per Capita ($)", labels=scales::dollar, breaks=seq(0,120000,20000))
p1
```

### Nonlinearities? Example: Quadratic Fit 

```{r, fig.height=4, echo=FALSE}
p1+stat_smooth(method = "lm", formula= y~poly(x,2), color="green")
```

### Nonlinearities? Example: No Outliers

```{r, fig.height=4, echo=FALSE}
p2<-gapminder %>%
  filter(gdpPercap<50000) %>% 
ggplot(., aes(gdpPercap,lifeExp))+
  geom_point(color="blue")+
  geom_smooth(method="lm", color="red")+
  scale_y_continuous(name="Life Expectancy (Years)", breaks=seq(0,90,20),limits=c(0,95))+
  scale_x_continuous(name="GDP per Capita ($)", labels=scales::dollar, breaks=seq(0,120000,20000))

p2
```

### Nonlinearities? Example: Quadratic Fit

```{r, fig.height=4, echo=FALSE}
p2+stat_smooth(method = "lm", formula= y~poly(x,2), color="green", size=2)
```

### Nonlinearities? Example: Logarithmic Fit

```{r, fig.height=4, echo=FALSE}
p2+stat_smooth(method = "lm", formula= y~poly(x,2), color="green", size=2)+
  stat_smooth(method = "lm", formula= y~log(x), color="orange", size=2)
```

### Nonlinear Regression

- Despite being "linear regression", OLS can handle this with an easy fix
- OLS requires all *parameters* (i.e. the $\beta$'s) to be linear, the *regressors* ($X$'s) can be nonlinear:

\vspace{-1cm}
\begin{align*}
		\onslide<3->{Y&=\beta_0+\beta_1 X^2 &  \checkmark \\}
		\onslide<4->{Y&=\beta_0+\beta_1^2 X &  x\\}
		\onslide<5->{Y&=\beta_0+\beta_1 \sqrt{X} & \checkmark\\}
		\onslide<6->{Y&=\beta_0+\sqrt{\beta_1} X  & x\\}
		\onslide<7->{Y&=\beta_0+\beta_1 X_1 * X_2 & \checkmark\\}
		\onslide<8->{Y&=\beta_0+\beta_1 ln(X) & \checkmark\\  }
\end{align*}
\vspace{-1cm}

\onslide<9->

- In the end, X will always be just a number, OLS can always estimate parameters for numbers
- *Plotting* the points $(X, \hat{Y})$ can result in a curve for nonlinear $X$'s

### Sources of Nonlinearities 

- The effect of $X \rightarrow Y$ may be nonlinear if:
1. $X \rightarrow Y$ is different for different levels of $X$
    - e.g. Diminishing returns: $\uparrow X$ increases $Y$ at a decreasing rate
    - e.g. Increasing returns: $\uparrow X$ increases $Y$ at an increasing rate
2. $X \rightarrow Y$ depends on the value of $X_2$
		- e.g. interaction terms (last lesson) 

### Sources of Nonlinearities II

\begin{columns}
		\begin{column}[c]{4.75cm}
			\begin{figure}
			\centering 
			\begin{tikzpicture}\scriptsize 
	\begin{axis}[
		scale=0.65,
		axis lines=middle, 
		enlarge x limits={rel=0.1, upper},
		enlarge y limits={rel=0.1, upper},
		every axis y label/.style={at={(axis description cs:-0.1,0.5)},rotate=90,anchor=north},
		every axis x label/.style={at={(axis description cs:0.5,-0.1)},anchor=north},
	%legend pos=outer north east,
	xlabel=X,
	ylabel=Y,
	shader=flat,
	ticks=none,
	grid=none,
	ymin=0,
	xmin=0,
	ymax=11,
]
	\addplot[domain=0:10, samples=20, ultra thick, color=blue!50]{10-x};
	\draw[very thick, dashed, red] (axis cs:3,7)--node[left]{$\Delta Y$}(axis cs:3,6)--node[below]{$\Delta X$}(axis cs:4,6);
\end{axis}
\end{tikzpicture}	
\caption*{\scriptsize Linear: slope $\beta_1=\frac{\Delta Y}{\Delta X}$ is constant} 
\end{figure}
		\end{column}
		\begin{column}[c]{4.75cm}
		\onslide<2->\begin{figure}
			\centering	
			\begin{tikzpicture}\scriptsize 
	\begin{axis}[
		scale=0.65,
		axis lines=middle, 
		enlarge x limits={rel=0.1, upper},
		enlarge y limits={rel=0.1, upper},
		every axis y label/.style={at={(axis description cs:-0.1,0.5)},rotate=90,anchor=north},
		every axis x label/.style={at={(axis description cs:0.5,-0.1)},anchor=north},
	%legend pos=outer north east,
	xlabel=X,
	ylabel=Y,
	shader=flat,
	ticks=none,
	grid=none,
	ymin=0,
	xmin=0,
	ymax=11,
]
	\addplot[domain=0:10, samples=20, ultra thick, color=blue!50]{10/(x)};
	\draw[very thick, dashed, red] (axis cs:1,9)--node[right]{$\Delta Y$}(axis cs:1,5)--node[below]{$\Delta X$}(axis cs:2,5);
	\draw[very thick, dashed, red] (axis cs:5,2)--node[left]{$\Delta Y$}(axis cs:5,1)--node[below]{$\Delta X$}(axis cs:9,1);
\end{axis}
\end{tikzpicture}	
\caption*{\scriptsize Quadratic: slope $\beta_1$ depends on value of $X$} 
\end{figure}
		\end{column}
		\begin{column}[c]{4.75cm}
			\onslide<3->\begin{figure}
			\centering 
			\begin{tikzpicture}\scriptsize 
	\begin{axis}[
		scale=0.65,
		axis lines=middle, 
		enlarge x limits={rel=0.1, upper},
		enlarge y limits={rel=0.1, upper},
		every axis y label/.style={at={(axis description cs:-0.1,0.5)},rotate=90,anchor=north},
		every axis x label/.style={at={(axis description cs:0.5,-0.1)},anchor=north},
	%legend pos=outer north east,
	xlabel=X,
	ylabel=Y,
	shader=flat,
	grid=none,
	ticks=none,
	xmin=0,
	ymax=11,
	ymin=0
]
	\addplot[domain=0:10, samples=20, ultra thick, color=blue!50]{2+x};
	\draw[very thick, dashed, red] (axis cs:2,4)--node[below]{\tiny $\Delta X$}(axis cs:4,4)--node[right]{\tiny $\Delta Y$}(axis cs:4,6);
	\draw[blue!50] (axis cs: 8, 8)node[below]{\tiny When $X_2=0$};
	\addplot[domain=0:10, samples=20, ultra thick, color=teal]{4+2*x};
	\draw[very thick, dashed, red] (axis cs:2,7.5)--node[below]{\tiny $\Delta X$}(axis cs:4,7.5)--node[right]{\tiny $\Delta Y$}(axis cs:4,11);
	\draw[teal] (axis cs: 6, 11)node[above]{\tiny When $X_2=1$};
\end{axis}
\end{tikzpicture}	
\caption*{\scriptsize Interaction: slope $\beta_1$ depends on value of $X_2$} 
\end{figure}
		\end{column}
	\end{columns}

# Polynomial Functions

### Polynomial Functions of $X$

$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2 + ... + \hat{\beta_{\textcolor{magenta}{r}}} X_i^{\textcolor{magenta}{r}} + \epsilon_i$$

- $r$ is the highest power $X_i$ is raised to (e.g. quadratic $r=2$, cubic $r=3$, etc)
    - The graph of an $r$^th^-degree polynomial function has $r-1$ bends 
- Just another multivariate OLS regression model  

# Quadratic Model

### Quadratic Model

$$\hat{Y_i} = \hat{\beta_0} + \hat{\beta_1} X_i + \hat{\beta_2} X_i^2 + \epsilon_i$$

- \alert{Quadratic model} has $X$ and $X^2$ variables in it (yes, need both!)
- How to interpret coefficients?
    - $\beta_0$ as "intercept" and $\beta_1$ as "slope" makes no sense 
    - $\beta_1$ as effect $X_i \rightarrow Y_i$ holding $X_i^2$ constant makes no sense
    - Note: this is *not* a multicollinearity problem! Correlation only measures *linear* relationships!
    - **Calculate marginal effects** by calculating predicted $\hat{Y_i}$ for different $X_i$

### Quadratic Model: Calculating Marginal Effects

- What is the effect of $\Delta X_i \rightarrow \Delta Y_i$? 
- Take the **derivative** of $Y_i$ with respect to $X_i$:
$$\frac{d Y_i}{d X_i} = \hat{\beta_1}+2\hat{\beta_2} X_i$$
- \alert{Marginal effect} of a 1 unit change in $X_i$ is a $\hat{\beta_1}+2\hat{\beta_2} X_i$ unit change in $Y$

### Quadratic Model: Example

\begin{example}
$$\widehat{\text{Life Expectancy}_i} = \hat{\beta_0}+\hat{\beta_1} GDP_i+\hat{\beta_2}GDP^2_i$$
	\begin{itemize}
		\item Life Expectancy
		\item GDP per Capita (GDP for short)
	\end{itemize}
\end{example}

### Quadratic Model: Simplifying

- These coefficients will be very large, let's first transform `gdpPercap` into \$1,000s, call it `gdp.t` [^1]

\onslide<2->
```{r}
gapminder <- gapminder %>%
  mutate(gdp.t=gdpPercap/1000)
```

\onslide<3->

- Let's also make the quadratic term by squaring `gdp.t` and calling it `gdp.sq`

\onslide<4->
```{r}
gapminder <- gapminder %>%
  mutate(gdp.sq=gdp.t^2)
```

[^1]: Note I am using `dplyr` and `%>%` here for efficiency, I loaded them before without showing it

### Quadratic Model: Example Regression in `R`

- Can "manually" run regression with `gdp.t` and squared term `gdp.sq`

\onslide<2->

\scriptsize 
```{r}
reg1<-lm(lifeExp~gdp.t+gdp.sq, data=gapminder)
summary(reg1)
```

### Quadratic Model: Example Regression in `R` II 

 - Or can use one of `R`'s shortcuts for polynomial models:
 - first term is normal, `x`, second term use `I(x^2)` to add squared term
 
\onslide<3->
\tiny

```{r}
reg1<-lm(lifeExp~gdp.t+I(gdp.t^2), data=gapminder)
summary(reg1)
```

### Quadratic Model: Example Regression in `R` III 

 - Another shortcut is just to use the `poly()` command
 - Instead of any `x` variables, just add `poly(x,2)` where 2 is the highest power desired[^2]
 
\onslide<3->
\tiny

```{r}
reg1<-lm(lifeExp~poly(gdp.t,2, raw=TRUE), data=gapminder)
reg1
```

[^2]: `R` gives different coefficient estimates for this version unless you add `, raw=TRUE` inside the `poly()` function, to ensure the polynomials are not computed orthogonally.

### Quadratic Model: Marginal Effects

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.02 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}	

- Marginal effect of GDP on Life Expectancy:

\begin{align*}
\only<2->{\frac{d \, Y}{d \; X} &= \hat{\beta_1}+2\hat{\beta_2} X_i\\}
\only<3->{\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55+2(-0.02)\text{GDP}\\}
\only<4->{ &= 1.55-0.04\text{GDP}\\}
\end{align*}

\onslide<5->

- Positive, with diminishing returns
- Effect on Life Expectancy of increasing GDP depends on initial value of GDP! 

### Quadratic Model: Marginal Effects II

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.02 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}	

- Marginal effect of GDP if GDP$=\$5$ (thousand): 

\begin{align*}
\only<2->{\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\}
		\onslide<3->{&= 1.55-0.04(5)\\}
		\onslide<4->{&= 1.55-0.20\\}
		\onslide<5->{&=1.35\\}
\end{align*}
\vspace{-1cm}
\onslide<6->

- i.e. for every addition \$1 (thousand) in GDP per capita, average life expectancy increases by 1.35 years

### Quadratic Model: Marginal Effects III

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.02 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}	

- Marginal effect of GDP if GDP$=\$25$ (thousand): 

\begin{align*}
\only<2->{\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\}
		\onslide<3->{&= 1.55-0.04(25)\\}
		\onslide<4->{&= 1.55-1.00\\}
		\onslide<5->{&=0.55\\}
\end{align*}
\vspace{-1cm}

\onslide<6->

- i.e. for every addition \$1 (thousand) in GDP per capita, average life expectancy increases by 0.55 years

### Quadratic Model: Marginal Effects IV

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.02 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}	

- Marginal effect of GDP if GDP$=\$50$ (thousand): 

\begin{align*}
\only<2->{\frac{d \, \text{Life Expectancy}}{d \, \text{GDP}} &= 1.55-0.04\text{GDP}\\}
		\onslide<3->{&= 1.55-0.04(50)\\}
		\onslide<4->{&= 1.55-2\\}
		\onslide<5->{&=-0.45\\}
\end{align*}
\vspace{-1cm}

\onslide<6->

- i.e. for every addition \$1 (thousand) in GDP per capita, average life expectancy *decreases* by 0.45 years

### Quadratic Model: Marginal Effects V

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.02 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}

\begin{table}	
	\centering 
	\begin{tabular}{rr}
	Initial GDP & Marginal Effect of +\$1,000 GDP\\ \toprule 
	\onslide<2->{\$5,000 & 1.35 years\\}
	\onslide<3->{\$25,000 & 0.55 years\\}
	\onslide<4->{\$50,000 & $-0.45$ years\\ \bottomrule }
	\end{tabular}
\end{table}

### Quadratic Model: Example

```{r, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(gdpPercap,lifeExp))+
  geom_point()+
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), color="green")+
  geom_vline(xintercept=5000, linetype=2, color="blue")+
  geom_vline(xintercept=25000, linetype=2, color="blue")+
  geom_vline(xintercept=50000, linetype=2, color="blue")+
  geom_text(x=5000, y=95, label="$5,000", color="blue")+
  geom_text(x=25000, y=95, label="$25,000", color="blue")+
  geom_text(x=50000, y=95, label="$50,000", color="blue")+
  scale_y_continuous(name="Life Expectancy (Years)", breaks=seq(0,90,20),limits=c(0,95))+
  scale_x_continuous(name="GDP per Capita ($)", labels=scales::dollar, breaks=seq(0,120000,20000))
```

### Quadratic Model: Maxima and Minima

- For a quadratic model, we can also find the predicted **maximum** or **minimum** of $\hat{Y_i}$
- By calculus, a minimum or maximum occurs where: 

\begin{align*}
		\onslide<3->{\frac{ d Y}{d X} &=0\\}
		\onslide<4->{\beta_1 + 2\beta_2 X &= 0\\}
		\onslide<5->{2\beta_2 X&= -\beta_1\\}
		\onslide<6->{X^*&=-\frac{1}{2}\frac{\beta_1}{\beta_2}\\}
		\end{align*}

### Quadratic Model: Maxima and Minima Example

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.015 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.00) & \\ 
\end{tabular}
\end{table}
	\begin{align*}
		\onslide<2->{X^*&=-\frac{1}{2}\frac{\beta_1}{\beta_2}\\}
		\onslide<3->{GDP^*&=-\frac{1}{2}\frac{(1.55)}{(-0.015)}\\}
		\onslide<4->{X^*&=-\frac{1}{2}(-103.333)\\}
		\onslide<5->{X^*&=51.67\\}
	\end{align*}

### Quadratic Model: Maxima and Minima Example II 

```{r, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(gdpPercap,lifeExp))+
  geom_point()+
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), color="green")+ # alternatively formula=y~poly(x,2)
  geom_vline(xintercept=51670, linetype=2, color="blue")+
  geom_text(x=51670, y=95, label="$51,670", color="blue")+
  scale_y_continuous(name="Life Expectancy (Years)", breaks=seq(0,90,20),limits=c(0,95))+
  scale_x_continuous(name="GDP per Capita ($)", labels=scales::dollar, breaks=seq(0,120000,20000))
```

# Determining if (Larger) Polynomials are Necessary

### Quadratic Model: Necessary?

\begin{table}[h!]
	\centering 
				\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccccc}
	$\widehat{\text{Life Expectancy}_i}=$ & 50.52 & $+$ & 1.55 & $GDP$ &$-$& 0.015 & $GDP^2$ \\
	 & \small (0.30) & & \small (0.04) & & & \small (0.0005) & \\ 
\end{tabular}
\end{table}

- Do we *need* a quadratic model? 
- We can determine if $\hat{\beta_2}$ is statistically significant:
    - $H_0$: $\beta_2=0$
    - $H_1$: $\beta_2\neq 0$

\onslide<5->
$$t=\frac{\hat{\beta_2}-0}{SE(\hat{\beta_2})}=\frac{0.015}{0.0005}=30$$
\onslide<6->

- Statistically significant $\implies$ we should keep the quadratic model
    - If we only ran a linear model, it would be biased!

### Higher-Order Polynomials: Cubic Regression

\begin{example}

\begin{itemize}
  \item Should we keep going up in polynomials? 
\end{itemize}

\onslide<2-> 
$$\widehat{\text{Life Expectancy}_i} = \hat{\beta_0}+\hat{\beta_1} GDP_i+\hat{\beta_2}GDP^2_i+\hat{\beta_3}GDP_i^3$$

\end{example}

\onslide<3->

```{r, echo=FALSE, fig.height=2.5}
ggplot(gapminder, aes(gdpPercap,lifeExp))+
  geom_point()+
  stat_smooth(method = "lm", formula = y ~poly(x,3), color="orange")+ 
  scale_y_continuous(name="Life Expectancy (Years)", breaks=seq(0,90,20),limits=c(0,95))+
  scale_x_continuous(name="GDP per Capita ($)", labels=scales::dollar, breaks=seq(0,120000,20000))
```

### Higher-Order Polynomials

- In general, should have a compelling theoretical reason why data or relationships should "change direction" multiple times 
- Or clear data patterns that have multiple "bends"

\onslide<2->
```{r, echo=FALSE, fig.height=3.5}
# make obviously cubic data

set.seed(1)
x <- seq(0,20,0.05)
y <- (500 + 0.4 * (x-10)^3)+rnorm(length(x), 10, 50) # real function + random noise
df<-data.frame(x,y)

ggplot(data=df, aes(x=x,y=y))+
  geom_point()+
  stat_smooth(method="lm", formula=y~poly(x,3), color="orange")+
  xlab("X")+ylab("Y")
```

### A Second Example

\begin{example}
	Test Scores: does school district's average income matter? 
\end{example}

```{r, echo=F, fig.height=3}
library("foreign")
CASchool<-read.dta("../Data/caschool.dta")

p3<-ggplot(CASchool, aes(avginc, testscr))+
  geom_point(color="blue")+
  xlab("District Average Income (Thousands of $)")+ylab("Test Score")+
  ylim(c(600,710))
p3
```

### A Second Example II

\begin{example}
	Test Scores: does school district's average income matter? 
\end{example}

$$TestScore_i=\hat{\beta_0}+\hat{\beta_1}Income_i$$

```{r, echo=F, fig.height=3}
library("foreign")
CASchool<-read.dta("../Data/caschool.dta")

p3+geom_smooth(method="lm", color="red")
```

### A Second Example III

\begin{example}
	Test Scores: does school district's average income matter? 
\end{example}

$$TestScore_i=\hat{\beta_0}+\hat{\beta_1}Income_i+\hat{\beta_2}Income^2_i$$	

```{r, echo=F, fig.height=3}
library("foreign")
CASchool<-read.dta("../Data/caschool.dta")

p3+geom_smooth(method="lm", color="red")+geom_smooth(method="lm", formula="y~poly(x,2)", color="green", size=2)
```

### A Second Example IV

- Let's manually generate a squared term, `avgincsq` 

\onslide<2->

```{r}
CASchool <- CASchool %>%
  mutate(avgincsq=avginc^2)
```

### A Second Example V

\columnsbegin

\column{0.5\textwidth}
\scriptsize 

```{r}
regsc<-lm(testscr~avginc+avgincsq, data=CASchool)
summary(regsc)
```

\column{0.5\textwidth}

\begin{table}
	\setlength{\tabcolsep}{0cm}
	\begin{tabular}{cccccccc}
	 $\widehat{\text{Test Score}}=$ & 607.30& $+$ & $3.85$ & $ Income$ & $-$ & $0.04$ & $Income^2$\\
	& \small (3.05) & & \small (0.30) & & & \small (0.01) & \\
	\end{tabular} 	
\end{table}

\columnsend

### A Second Example VI

\begin{example}
	Test Scores: does school district's average income matter? 
\end{example}

$$TestScore_i=\hat{\beta_0}+\hat{\beta_1}Income_i+\hat{\beta_2}Income^2_i+\hat{\beta_3}Income^3_i$$

```{r, echo=F, fig.height=3}
p3+geom_smooth(method="lm", color="red")+
  geom_smooth(method="lm", formula="y~poly(x,2)", color="green", size=2)+
  geom_smooth(method="lm", formula="y~poly(x,3)", color="orange", size=2)
```

### A Second Example VII

- Let's manually generate a cubic term, `avginc3`: 

\onslide<2->

```{r}
CASchool <- CASchool %>%
  mutate(avginc3=avginc^3)
```

### A Second Example VIII

\scriptsize 

```{r}
regcu<-lm(testscr~avginc+avgincsq+avginc3, data=CASchool)
summary(regcu)
```

### Strategy for Polynomial Model Specification

1. Are there good theoretical reasons for relationships changing (e.g. increasing/decreasing returns)?
2. Plot your data: does a straight line fit well enough?
3. Specify a polynomial function of a higher power (start with 2) and estimate OLS regression
4. Use $t$-test to determine if higher-power term is significant
5. Interpret effect of change in $X$ on $Y$ 
6. Repeat steps 3-5 as necessary

```{r, echo=F, eval=F}
func<-function(x){x}

ggplot(data.frame(x=0:10),aes(x=x))+
    stat_function(fun=log, color="blue")+
    stat_function(fun=exp, color="red")+
  stat_function(fun=func, linetype=2)+
  ylim(c(0,10))
```