---
title: "A Modest Handbook for R"
author: "Ryan Safner"
date: "9/28/2018"
output: 
  html_document:
    theme: simplex
    toc: true 
    highlight: tango
---

> "There’s an implied contract between you and R: it will do the tedious computation for you, but in return, you must be completely precise in your instructions. Typos matter. Case matters." - Hadley Wickham, [R for Data Science](http://r4ds.had.co.nz/workflow-basics.html)

# Basic Concepts

# Hints for Writing Code

## Calculations

- `R` can be used as a calculator 
    - Basic operations `+`, `-`, `*`, `/`
    - More advanced math operators like exponents, logarithms, trigonometric functions, etc

```{r}
2+2
6^2 # 6 to the second power (i.e. squared)
sqrt(100/4) # square root 
log(5) # logarithm
sin(2*pi) # sin
factorial(5) # factorial (e.g. 5!)
choose(2,6) # binomial choose function

# order of operations matters
3*3+4
3*(3+4)
```

- `R` often reports very large (or very small) numbers in scientific notation with `e`
    - For positive `e`: the number of zeros (or digits after the decimal point) to the right of a number 
        - e.g. $1.25e6 = 1.25 \times 10^6 = 1,250,000$
    - For negative `e`: one less than the number of zeros (or digits after the decimal point) to the left of a number 
        - e.g. $1.25e-6 = 1.25 \times 10^{-6} = 0.00000125$
        
# Objects

## Vector 

- The simplest data structure in `R` is a \alert{vector}, simply a collection of objects
- To construct a vector, use the "combine/concatenate" function "**c()**"
- **Example**: a vector of the numbers 1 through 5

```{r vector1}
c(1,2,3,4,5)
```

- We can also build vectors via generating series

\onslide<6->
```{r vector 11}
1:5
```


- We can perform mathematical operations on a vector as a whole:
```{r vector 3}
sum(1:5)
mean(1:5)
```

- `R` is an \alert{object-oriented} programming language 
- We will almost always store data in **objects** and run **functions** on objects
    - Objects are assigned values: `objectname<-value`
    - Functions take the form: `functionname(objectname)`
- Storing a vector as an object called "*x*" using the assignment operator "**<-**" 

```{r store.vector.as.x}
x<-c(1,2,3,4,5)
```
- To inspect an object, we simply call it up by name

```{r look.at.x}
x
```

# Data Classes


- Vectors **must** contain the same type of elements (e.g. numerical or text)
  * Technically this refers to **atomic vectors** (nearly all vectors are atomic)
- Vectors with "mixed" types will convert all elements to the lowest-common denominator, e.g. character
- You can always check the type of vector using **class()**
```{r vector.types}
mixed<-c(5, pi, TRUE, 4.3, "cabbage")
class(mixed)
```

### Numeric

- \alert{Numeric} (aka "double"), as it sounds, can perform mathematical operations
```{r numeric, echo=TRUE, eval=FALSE}
numeric<-c(1,2,3,4,5)
```

- There are two **types** of numeric objects: **double** and **integer** 

### Double

- If numeric values contain decimal points, they are technically called \alert{floating point double} or simply \alert{double} class
- `R` may simply call them numeric, but contrast with integer below 

```{r}
double<-c(pi,2.34,9.99)

class(double)
typeof(double) # will return the more specific type 

is.double(double) # a logical test to see if object is "double" type
is.integer(double) # a logical test to see if object is "integer" type
```

#### Integer

- If numeric values are all whole numbers, they are \alert{integer} class

```{r}
integers<-c(1,2,3,4)
class(integers)
typeof(integers)

is.double(double)
is.integer(double)
```

### Logical

- \alert{Logical} is a series of binary elements or statements that can either be TRUE or FALSE

```{r logical, echo=TRUE,eval=FALSE}
logical<-c(TRUE,FALSE,FALSE,TRUE)
```

- We can perform logical tests with common operators:
    - `<` less than
    - `>` greater than
    - `<=` less than or equal to ($\leq$)
    - `>=` greater than or equal to ($\geq$)
    - `==` is equal to (note two equals signs are needed!)
    - `!=` is not equal to 
    - `%in%` is a member of a set ($\in$)

```{r}
3==4 #is 3 equal to 4?
3<4 # is three less than 4?
3<=4 # is three less than or equal to 4?
3>4 # is three greater than 4?
3!=4 # is three not equal to four?
3 %in% c(0,1,2) # is three in the following set of numbers?
3 %in% c(0,1,2,3) # is three in the following set of numbers?
```

- We are not limited to using numeric data, `R` can also perform logical tests on other classes of variable, like characters (which need quotes): 

```{r}
"red"=="blue" # is red the same as blue?
"red"!="blue" # is red not equal to blue?
political.party<-c("Republican","Democrat") # define political party as a set of Republican and Democrat
"Libertarian" %in% political.party # check if Libertarian is in the set of political parties we created
"Democrat" %in% political.party # check if Democrat is in our set of political parties 
```

- We can also perform more than one test at a time with multiple conditions:
    - `&` AND 
    - `|` OR 

```{r}
2==2 & 2>3 # is 2 equal to 2 AND greater than 3? 
2==2 | 2>3 # is 2 equal to 2 OR greater than 3? 
```

- These commands will become very useful when we want to subset data or look at portions of our data based on some condition

### Character

- \alert{Character} is a string of text: letters, numbers, and symbols, cannot perform mathematical operations
    - Character values require quotation marks around each value 
```{r char, echo=TRUE,eval=FALSE}
character<-c("one","two","7","orange")
```

#### Dates

- Dates are a specific type of character class
- Specific dates
    * Can do days, weeks, months, quarters, years
```{r dates 1}
today<-Sys.Date() #print today's date 
format(today, format="%B %d %Y") # specify how to report date format 

months<-seq(as.Date("2010/1/1"), as.Date("2012/1/1"), "months") # generate sequence of months between Jan 2010 and Jan 2012
months
```

### Factor 

- \alert{Factor} is a special type of character variable, often used to indicate membership in one of several possible categories, called **levels** (e.g. for plotting, or conditional statistics and data work)
```{r factors}
students<-factor(c("freshman", "senior", "senior", "junior", "freshman", 
                   "sophomore", "freshman"))
students # note order is arbitrary
levels(students) #extract unique levels 
nlevels(students) #count the number of levels 
```

```{r factors2}
table(students) #tabulate number of values for each level 
```

#### Ordered Factors

```{r}
students.o<-ordered(students, levels=c("freshman","sophomore","junior","senior"))
students.o
```

- **Be advised**: when `R` stores and calls factors, it actually stores them as integers [1..k, for k categories] instead of characters (e.g. "freshman"=1, "sophomore"=2), making this a **nominal** variable. This allows for some mathematical operations.
- An **ordered factor** is where the ordering matters (e.g. "small", "medium", "large" coded as 1, 2, 3 in order)

## Checking or Reclassifying Objects

- We can always check the class of an object with `class()` or `typeof()`. 
    - We can perform logical tests `is.numeric()`, `is.factor()`, etc. to see if an object is a specified type
    - We can change the class of an object by redefining it with `as.classname()`, e.g.

```{r class change}
is.numeric(x) # check if x is numeric
is.factor(x) # check if x is a character
x<-as.character(x) # change vector x to a character
class(x)
x<-as.numeric(x) # change vector x back to numeric 
class(x)
```

# Other Object Types

## Lists 

- A \alert{list} is a non-atomic vector, meaning you can gather data elements of different classes in one object 

```{r list}
mylist<-list(5, pi, TRUE, 4.3, "cabbage")
class(mylist)
```

- Another great property of lists is that elements of the list can themselves be vectors

```{r}
vectored.list<-list(c(1.82, 1940, 93.20, 192.917),
     c("Orange", "Cyan", "Pink"),
     c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE))

str(vectored.list) # look at structure of the list
```

- We can create a **label** for each element in a list, called a `name` 

```{r}
vectored.list<-list(numbers=c(1.82, 1940, 93.20, 192.917), # first element is a vector called 'numbers'
     colors=c("Orange", "Cyan", "Pink"), # second element is a vector called `colors`
     logic=c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE)) # third element is a vector called `logic`

vectored.list
```

- The `names` command prints (or changes) the name of the label of each element in the list

```{r}
names(vectored.list) # print the names of the list elements
names(vectored.list)<-c("name1","name2","name3") # rename the lables to 'name1', 'name2', and 'name3'

names(vectored.list) # print new names
vectored.list # print list with new names
```

## Matrix

- Everything thus far has been 1 dimension, but we often work with 2-dimensional data
    - Rows are observations
    - Columns are variables
- A \alert{matrix} 
    - `matrix()` command creates a matrix by column,
        - can define number of rows with `nrow=`, `R` will divide the elements into equal number of columns 
    
```{r}
matrix1<-matrix(c(1,2,3,4,5,6),nrow=3) # make a 3-row matrix
matrix1 
```

## Data Frame

- The most important object in `R` is a **data frame** (what you call a "spreadsheet"), used for statistics, plots, regressions, etc
    - "Rectangular" data, rows are observations, columns are variables 
    - Can hold variables of different classes (e.g. a quantitative variable like income, a character variable like name, etc)
    - In essence, data frames are actually lists (where each list object itself is a vector)
    - All vectors (columns) must have the same length! 

```{r}

df<-data.frame(x = 1:3,
           y = c("a", "b", "c"),
           z = c(TRUE, FALSE, TRUE))
df
```

# Manipulating Objects

## Indexing and Subsetting Objects

- `R` objects store multiple values as an ordered list
    - The order of each value is kept as an index number
        - e.g. the first value is [1], the second value is [2], the 82nd value is [82], etc. 

- Use [square brackets] to isolate elements of a vector, index starts at [1]

```{r vector.index}
print(mylist) #Our original vector 
mylist[1] #Inspect first element 
mylist[c(1,3)] #Inspect first and third elements
mylist[2]<-"orange" #Change second element
print(mylist) #Our new vector
```

- We can also ask how many elements a vector contains with **length()**
```{r length}
length(list)
```

- We can also include expressions when requesting an indexed value 
```{r length.2}
#Get second-to-last value in list (length of list minus 1 value)
mylist[length(mylist)-1]
```

- It is often useful to label values (later, variables) by giving them "names"
```{r names}
myvars<-c(Var1="alpha",Var2=12,Var3="height") #Label name goes first, then =, then the value 
myvars

#We can still call specific values by index position 
myvars[3] #Get 3rd value 

#We can equivalently call up specific values by name 
myvars["Var3"] #Get value of Var3
names(myvars) #List names of all values
names(myvars)[3] #Get name of third value
```

- 


## Math Operations on Objects

## Redefining Objects

# Plotting 

## Base `R`

- Base `R` is very powerful and intuitive to plot, but not very sexy 
- Basic syntax: 

```{r, eval=FALSE}
plottype(data.frame.name$variable)
```

- If using multiple variables, you can avoid `$` by typing the variable names and then telling `R` where the variables come from (a data frame)

```{r, eval=FALSE}
plottype(variable1,variable2, data=data.frame.name)
```

### Histogram

```{r, echo=TRUE, fig.height=4, fig.width=10, warning=FALSE}
library("gapminder")
hist(gapminder$gdpPercap)
```

### Boxplot 

- Boxplots are similar syntax

```{r, fig.height=4}
boxplot(gapminder$gdpPercap)
```

- If we want a boxplot for each category, use `variable.name~category.variable.name` to tell `R` to plot a boxplot **by** category 

```{r, fig.height=4}
boxplot(gdpPercap~continent,data=gapminder)
```

### Scatterplot 

- Scatterplot syntax for plotting is similar to `hist()` and `boxplot()`: `plot(df$x,df$y)`
```{r, fig.height=3.5}
plot(gapminder$gdpPercap, gapminder$lifeExp)
```

## With `ggplot2`

# Linear Regression

```{r}
sex<-c("Male", "Female")
marital<-c("Married","Unmarried")
```

