---
title: "Interpretting Regression Coefficients"
author: "Ryan Safner"
date: "12/4/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning=F, message=F)
```

How we interpret the coefficients in regression models will depend on how the dependent ($Y$) and independent ($X$) variables are measured. In general, there tend to be three types of variables used in econometrics: continuous variables, the natural log (ln) of continuous variables, and dummy variables. In the examples below, we will consider models with three different independent variables:

- $X_{1i}$: a continuous variable
- $ln(X_{2i})$: the natural log of a continuous variable
- $X_{3i}$: a dummy variable that equals 1 (if yes) or 0 (if no)

Below are three different OLS models. In each case, we keep the right hand side variables are the same, but as a demonstration, we change the dependent variable ($Y$) of interest to show the difference when we measure it as a continuous variable, the natural log of a continuous variable, or a dummy variable:

- $Y_{1i}$: a continuous variable
- $ln(Y_{2i})$: the natural log of a continuous variable
- $Y_{3i}$: a dummy variable that equals 1 (if yes) or 0 (if no)

## Model 1

$$Y_{1i}=\beta_0+\beta_1X_{1i}+\beta_2ln(X_{2i})+\beta_3X_{3i}+\epsilon_i$$

- $\beta_1=\frac{\Delta Y_{1i}}{\Delta X_{1i}}$: a one unit change in $X_1$ causes a $\beta_1$ unit change in $Y_{1i}$
- $\beta_2=\frac{\Delta Y_{1i}}{\Delta ln(X_{2i})}$: a 1\% change in $X_2$ causes a $0.01\times\beta_2$ unit change in $Y_{1i}$
- $\beta_3=\frac{\Delta Y_{1i}}{\Delta X_{3i}}$: the change in $X_3$ from 0 to 1 causes a $\beta_3$ unit change in $Y_{1i}$	

## Model 2

$$ln(Y_{2i})=\beta_0+\beta_1X_{1i}+\beta_2ln(X_{2i})+\beta_3X_{3i}+\epsilon_i$$

- $\beta_1=\frac{\Delta ln(Y_{2i})}{\Delta X_{1i}}$: a one unit change in $X_1$ causes a $100\times\beta_1$ percent change in $Y_{2i}$
- $\beta_2=\frac{\Delta ln(Y_{2i})}{\Delta ln(X_{2i})}$: a 1\% change in $X_2$ causes a $\beta_2$ percent change in $Y_{2i}$
- $\beta_3=\frac{\Delta Y_{1i}}{\Delta X_{3i}}$: the change in $X_3$ from 0 to 1 causes a $100\times\beta_3$ percent change in $Y_{2i}$	

## Model 3

$$Y_{3i}=\beta_0+\beta_1X_{1i}+\beta_2ln(X_{2i})+\beta_3X_{3i}+\epsilon_i$$

- $\beta_1=\frac{\Delta Y_{3i}}{\Delta X_{1i}}$: a one unit change in $X_1$ causes a $100\times\beta_1$ percentage point change in the probability of $Y_{3i}$ occurring (=1)
- $\beta_2=\frac{\Delta Y_{3i}}{\Delta ln(X_{2i})}$: a 1\% change in $X_2$ causes a $\beta_2$ percentage point change in the probability of $Y_{3i}$ occurring (=1)
- $\beta_3=\frac{\Delta Y_{3i}}{\Delta X_{3i}}$: the change in $X_3$ from 0 to 1 causes a $100\times\beta_3$ percentage point change in the probability of $Y_{3i}$ occurring (=1)

# Example With Data


Below are the results from three regressions using the same data set. The results parallel the three general models outlined above. The dataset `meps2005.dta` can be found under Blackboard/Datasets. It contains responses from a sample of senior citizens all on Medicare. 

The regressions have three different outcome measures (analogous to $Y_1$, $Y_2,$ and $Y_3$ above): total expenditures on medical care (`totalexp`, $Y_1$), the natural log of total expenditures on medical care (`ln.totalexp`, $Y_2$), and whether or not the person reports "`goodhealth`" ($Y_3$). 

For each of these three dependent variables, we regress three potential independent variables, a continuous variable (`age`), the natural log of a continuous variable (`ln.income`), and a dummy variable (`obese`$=$1 if a person is obese, $=0$ otherwise). The sample description and summary statistics are presented below: 

```{r, load-data, echo=F}
library("readstata13")
handout<-read.dta13("../Data/meps_2005.dta")

```

```{r, summary-stats, echo=F}
suppressPackageStartupMessages(library("dplyr"))
library("tidyr")
library("knitr")
handout<- handout %>%
  rename(ln.totalexp="ln_totalexp",
         ln.income="ln_income")
```

```{R}
handout.sum<-handout %>%
  select(totalexp, ln.totalexp, goodhealth, age, ln.income, obese) %>% 
  summarize_each(funs("Obs"=n(),
            "Mean"=round(mean(.,na.rm=TRUE),2),
            "Std. Dev."=round(sd(., na.rm=TRUE),2),
            "Min"=round(min(., na.rm=TRUE),2),
            "Q1"=round(quantile(., 0.25, na.rm=TRUE),2),
            "Median"=round(median(., na.rm=TRUE),2),
            "Q3"=round(quantile(., 0.75, na.rm=TRUE),2),
            "Max"=round(max(., na.rm=TRUE),2)
            ))

handout.sum.tidy<- handout.sum %>%
  gather(key=stat,value=val) %>%
  separate(stat, into=c("var", "stat"), sep="_") %>% 
  spread(stat, val) %>% 
  select("var", "Obs", "Min", "Q1", "Median", "Q3", "Max", "Mean", "Std. Dev.") %>%
  rename(Variable="var") %>% 
  kable(.) 

handout.sum.tidy
```

\clearpage 

# Model 1

$$\widehat{Totalexp}=\hat{\beta_0}+\hat{\beta_1}age+\hat{\beta_2}ln(income)+\hat{\beta_3}obese$$

```{r, echo=F}
lm(totalexp~age+ln.income+obese, data=handout) %>%
  summary()
```

$$\widehat{Totalexp}=-6857.36+194.08age+44.30ln(income)+1393.60obese$$

Interpreting the coefficients: 

- `age`: a one year increase in age will increase annual medical expenditures by $194
- `ln.income`: a 1\% increase in income will increase medical spending by $0.01\times44.2=\$0.442$
- `obses`: obese seniors spend $1,393 more per year on medical care than non-obese seniors

\clearpage 

# Model 2

$$\widehat{ln(Totalexp)}=\hat{\beta_0}+\hat{\beta_1}age+\hat{\beta_2}ln(income)+\hat{\beta_3}obese$$

```{r, echo=F}
lm(ln.totalexp~age+ln.income+obese, data=handout) %>%
  summary()
```

$$\widehat{ln(Totalexp)}=6.17+0.044age-0.16ln(income)+0.45obese$$

Interpreting the coefficients:

- `age`: a one year increase in age will increase annual medical expenditures by 4.37\%
- `ln.income`: a 1\% increase in income will reduce medical spending by 0.16\%
- `obese`: obese seniors spend 44.6\% more per year on medical care than non-obese seniors

\clearpage 

# Model 3

$$\widehat{Goodhealth}=\hat{\beta_0}+\hat{\beta_1}age+\hat{\beta_2}ln(income)+\hat{\beta_3}obese$$

```{r, echo=F}
lm(goodhealth~age+ln.income+obese, data=handout) %>%
  summary()
```

$$\widehat{Goodhealth}=-0.421+0.003age+0.079ln(income)+0.167obese$$

Interpreting the coefficients:

- `age`: a one year increase in age will increase the probability of reporting good health by 0.3 percentage points
- `ln.income`: a 1\% increase in income will increase the probability of reporting good health by 0.079 percentage points
- `obese`: obese seniors have 16.7 higher percentage point probability of reporting good health than non-obese seniors
