---
title: "Understanding `factor`s in `R` Practice"
author: "Ryan Safner"
date: "11/19/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's generate some random data. We will make a `data.frame` called `df` (feel free to call it something else.)
```{r}
set.seed(1) # this makes data reproducible, if we set the same seed #, we all get the same "random" data

colors<-c("red", "orange","yellow", "green", "blue", "purple")
color<-sample(colors,100,replace=TRUE)

educ.levels<-c("high school","college", "graduate degree")
education<-sample(educ.levels, 100, replace=TRUE)


# make a data frame of three variables, x, y, and color
df<-data.frame(x=rnorm(100,2,1), # x is 100 draws from a normal distr with mean 2 and sd 1
                 y=rnorm(100,5,1), # y is 100 draws from a normal distr with mean 5 and sd 1 
                 color=factor(color), # color is a factor
                 education=ordered(education, levels=c("high school", "college", "graduate degree")) # education is an ordered factor, specify the proper ranking from fist to last
               ) 
```

### 1. Get a summary of `df`. Also check its `str()` and `head()` to get a closer look. Look at the data itself with `View(df)`. 

```{r}
summary(df) # get summary of data
str(df)
head(df)
#View(df)
```

### 2. Look more closely at `color`. Check its `class()`, `nlevels()` and the actual `levels()`. Finally, make a `table()` of the counts of each category. 

```{r}
class(df$color)
nlevels(df$color)
levels(df$color)
table(df$color) # get counts of color 
```

### 3. Make a barplot of `color`. 

```{r, fig.height=2.5}
library(ggplot2)

ggplot(data = df, aes(x = color))+
  geom_bar(aes(fill=color))
```

### 4. Look more closely at `education`. Check its `class()`, `nlevels()` and the actual `levels()`. Finally, make a `table()` of the counts of each category. 

```{r}
class(df$education)
nlevels(df$education)
levels(df$education)
table(df$education) # get counts of color 
```

### 5. Make a barplot of `education`. 

```{r, fig.height=2.5}
ggplot(data = df, aes(x = education))+
  geom_bar(aes(fill=education))
```

### 6. Now let's try looking at plots by different categories. Make a scatterplot of `x` and `y` and add `color=color` to your base layer `aes()`.

```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y, color=color))+
  geom_point()
```

#### a. Now let's try subsetting. Plot only data for the color `green`.  

```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y))+
  geom_point(aes(color=color))+
  geom_smooth(method="lm")
```

#### b. In addition to your `geom_point()`, add a `geom_smooth(method="lm")` regression line. Notice it makes a regression line for each color. If we want an overall regression line, we need to redo our scatterplot as follows. In the base layer, don't include `color` in your `aes()`, move it instead inside `geom_point(aes(color=color))`. Then add a `geom_smooth()`, it will do it for the overall plot. 

```{r, fig.height=2.5}

ggplot(data = df, aes(x = x, y = y))+
  geom_point(aes(color=color))+
  geom_smooth(method="lm")
```

#### c. Now let's try subsetting. Make a scatterplot and regression line only with data points for the color `green`.  

```{r, fig.height=2.5}
# two ways to do this, one using the index [row, columns] selecting only rows for which color is green

ggplot(data = df[df$color=="green",], aes(x = x, y = y))+
  geom_point()+
  geom_smooth(method="lm")

```

```{r, fig.height=2.5}
# another is to use subset and make another df to use 

df.green<-subset(df,color=="green")

ggplot(data = df.green, aes(x = x, y = y))+
  geom_point()+
  geom_smooth(method="lm")
```

#### d. Let's simply use the `facet_grid()` command to plot all the different colors as different plots. Reuse your commands from the first plot in this question, and then add a `facet` layer with `+facet_grid(cols=vars(color))`

```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y))+
  geom_point(aes(color=color))+
  geom_smooth(method="lm")+
  facet_grid(cols=vars(color))
```

### 7. Run through problem \#6 again, but using `education` instead of `color`.

```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y, color=education))+
  geom_point()
```


```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y))+
  geom_point(aes(color=education))+
  geom_smooth(method="lm")
```

```{r, fig.height=2.5}
df.hs<-subset(df,education=="high school")

ggplot(data = df.hs, aes(x = x, y = y))+
  geom_point()+
  geom_smooth(method="lm")
```

```{r, fig.height=2.5}
ggplot(data = df, aes(x = x, y = y))+
  geom_point(aes(color=education))+
  geom_smooth(method="lm")+
  facet_grid(cols=vars(education))
```

### 8. Now let's try some regression. 

### a. Run a regression of `y` on `x` and `education`. What happens? 

```{r}
reg<-lm(y~x+education, data=df)
summary(reg)
```

### b. `R` was generous and did the work for you! But let's do the same thing ourselves manually. What we need to do is convert `education` into a three dummy variables, one for each level of education. It's easiest to use the `ifelse()` command here. Remember the syntax: `ifelse(condition, do.this.if.true, do.this.if.false)`. Check your data `df` again with `head()` or `View()` to make sure you properly coded the variables. 

```{r}
df$hs<-ifelse(education=="high school",1,0)
df$college<-ifelse(education=="college",1,0)
df$grad<-ifelse(education=="graduate degree",1,0)
```

```{r}
head(df)
```

### c. Now run a regression of `y` on `x` and **all** of your new dummy variables. What happens, and why?

```{r}
reg2<-lm(y~x+hs+college+grad, data=df)
summary(reg2)
```

### d. Run three different regressions, each one omitting one of the different categories of education. Interpret your coefficients.

```{r, results="asis"}
reg.no.hs<-lm(y~x+college+grad, data=df)
reg.no.college<-lm(y~x+hs+grad, data=df)
reg.no.grad<-lm(y~x+hs+college, data=df)

suppressPackageStartupMessages(library("stargazer"))
stargazer(reg.no.hs, reg.no.college, reg.no.grad, type="latex", float=F, header=F)
```
