---
title: "R Practice 2: `ggplot2` and Simple Regression"
author: "Ryan Safner"
date: "October 3, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Install and load the package `gapminder`. Type `?gapminder` and hit enter to see a description of the data. 

```{r}
# install.packages("gapminder") #uncomment for initial installation
library("gapminder") # load gapminder
?gapminder 
```

### 2. Get summary statistics of `gpdPercap`. 

```{r}
summary(gapminder$gdpPercap)
```

### 3. Use base `R`'s `hist()` function to plot a histogram of `gdpPercap`

```{r, fig.height=3.5}
hist(gapminder$gdpPercap)
```

### 4. Now load and use `ggplot2` to create a histogram of `gdpPercap`. Remember your base layer must establish which `data frame` you are using (`gapminder`) and the base aesthetics `aes()` to define what variable is `x`. Your second layer is a `geom_histogram()`

```{r, fig.height=3.5}
library("ggplot2") # load ggplot2

ggplot(gapminder,aes(x=gdpPercap))+
  geom_histogram()
```

### 5. Get summary statistics of `lifeExp`.

```{r}
summary(gapminder$lifeExp)
```

### 6. Use base `R`'s `hist()` function to create a histogram of `lifeExp`.

```{r,fig.height=3.5}
hist(gapminder$lifeExp)
```

### 7. Use `ggplot2` to create a histogram of `lifeExp`. 

```{r,fig.height=3.5}
ggplot(gapminder,aes(x=lifeExp))+
    geom_histogram()
```

### 8. Instead of a histogram, make a density plot of `lifeExp` with `geom_density()`

```{r,fig.height=3.5}
ggplot(gapminder, aes(x=lifeExp))+
  geom_density()
```

### 9. Using base `R`'s `boxplot()` function, create a boxplot of `gpdPercap` by `continent`. 

```{r,fig.height=3.5}
boxplot(gdpPercap~continent, data=gapminder)
```

### 10. Now do the same with `ggplot2`. In your initial aesthetics, set `x` as `continent`, `y` as `gdpPercap` and `fill` (color) by `continent`. Your `geom` layer is `geom_boxplot()`. 

```{r,fig.height=3.5}
ggplot(gapminder, aes(x=continent, y=gdpPercap, fill=continent))+
  geom_boxplot()
```

### 11. The nice thing about building plots one layer at a time is that we can use different `geoms` on the same base layer. Replicate your answer to \#10 and instead of `geom_boxplot()`, try a "Violin plot" with `geom_violin()`. 

```{r,fig.height=3.5}
ggplot(gapminder, aes(x=continent, y=gdpPercap, fill=continent))+
  geom_violin()
```

### 12. Use what you've learned so far to make a density plot of `gdpPercap` by `continent`. Note your only variable here is `x`. Add an option to your `geom_density` layer of setting `alpha=0.5` (to make plots more transparent). 

```{r,fig.height=3.5}
ggplot(gapminder, aes(x=gdpPercap, fill=continent))+
  geom_density(alpha=0.5)
```

### 13. Do the same thing for `lifeExp`

```{r,fig.height=3.5}
ggplot(gapminder, aes(x=lifeExp, fill=continent))+
  geom_density(alpha=0.5)
```

### 14. Now let's try to estimate the following relationship. 

$$\widehat{\text{Life Expectancy}}=\beta_0+\beta_1 \text{GDP Per Capita}$$

First, use base `R` to make a scatterplot of these two variables with `plot()`. Be sure to signify `x` and `y` using the `data.frame$variable` syntax. 

```{r,fig.height=3.5}
plot(gapminder$gdpPercap,gapminder$lifeExp)
```

### 15. Now let's try with `ggplot2`. For your base layer, consider in your aesthetics what is `x` and what is `y`. We want our data to manifest as data points, so use `geom_point()` as your second layer. Be sure to save this as some object. 

```{r,fig.height=3.5}
# my object is called p 
p<-ggplot(gapminder,aes(x=gdpPercap,y=lifeExp))+
  geom_point()
p
```

### 16. Now on top of the existing plot, let's add a regression line. Redefine your object to be itself `+geom_smooth(method="lm")` to add the regression line (`geom_smooth` creates a smooth line, and `lm` stands for linear model, i.e. OLS regression).

```{r,fig.height=3.5}
p<-p+geom_smooth(method="lm")
p
```

### 17. Now let's spice this up a bit. Recreate your plot but this time, include in your base layer's aesthetics (in addition to defining `x` and `y`)  `color=continent` to color by continent. 

```{r,fig.height=3.5}
p<-ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent))+
  geom_point()
p
```

### 18. Now add a regression line. Notice that since we initially defined in the base layer to color by continent, it also creates different colored lines, one for each continent.  

```{r,fig.height=3.5}
p<-ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent))+
  geom_point()+geom_smooth(method="lm")
p
```

### 19. Let's try facetting. Add to your previous plot `+facet_grid(cols=vars(continent))`. This creates a grid of individual plots, one for each continent, and arranges them into columns (`cols`) by the variable `continent`. 

```{r,fig.height=3.5}
p<-ggplot(gapminder,aes(x=gdpPercap,y=lifeExp,color=continent))+
  geom_point()+geom_smooth(method="lm",color="black")+facet_grid(cols=vars(continent))
p
```

### 20. Let's try only looking at the year 2002. We can use the `subset()` function to create another data frame for only the year 2000 like `gapminder.2002<-subset(gapminder, year==2002)`. Next, get summary statistics for the gdp per capita in 2002. 

```{r,}
gapminder.2002<-subset(gapminder, year==2002)
summary(gapminder.2002$gdpPercap)
```

### 21. Plot a histogram of gdp per capita in `ggplot2` for 2002 

```{r,fig.height=3.5}
ggplot(gapminder.2002,aes(x=gdpPercap))+
  geom_histogram()
```

### 22. Plot a scatterplot with `ggplot2` for 2002 gdp per capita (`x`) vs. life expectancy (`y`)

```{r,fig.height=3.5}
ggplot(gapminder.2002,aes(x=gdpPercap,y=lifeExp))+
  geom_point()+geom_smooth(method="lm")
```

### 23. Now let's add more information to our scatterplot. Add an option to the `geom_point()` to plot `size=pop`. 

```{r,fig.height=3.5}
ggplot(gapminder.2002,aes(x=gdpPercap,y=lifeExp, color=continent))+
  geom_point(aes(size=pop))
```

\clearpage

# Regression Analysis

### 24. Now let's turn away from data visualization to more technical analysis with regression. For more information and examples, see lecture 7. Run a regression of life expectancy on gdp per capita. `summary()` your regression. What are: 

- $\hat{\beta_0}$
- $\hat{\beta_1}$
- $SE(\hat{\beta_0})$
- $SE(\hat{\beta_1})$
- $R^2$
- $SER$

```{r}
reg<-lm(lifeExp~gdpPercap,data=gapminder)
summary(reg)
```

- $\hat{\beta_0}=53.96$
- $\hat{\beta_1}=0.0007649$ (for every 1 \$ increase in GDP, life expectancy increases by 0.0007649 years)
- $SE(\hat{\beta_0})=0.315$
- $SE(\hat{\beta_1})=0.00002579$
- $R^2=0.3407$ (our model explains 34\% of the total variation in Life Expectancy)
- $SER=10.49$ (the average prediction is off by 10.49 years of Life Expectancy)

### 25. Is $\hat{\beta_1}$ statistically significantly different from 0 (i.e. $H_0: \beta_1=0, H_1: \beta_1 \neq 0$)? How do you know? See lecture 8 for more help. 

Yes. The $t$-statistic is 29.66, which is very large, and the $p$-value is 0.000000000000002, which is very small. 

We also see that the estimated slope is more than twice its' standard error: $0.0007649>2(0.00002579)$

### 26. Save the residuals and plot them in a residual plot (using the residuals as `y` instead of `lifeExp`). Add a horizontal line at 0 with `geom_vline(yintercept=0)`

```{r,fig.height=3.5}
gapminder$res<-residuals(reg)

ggplot(gapminder,aes(x=gdpPercap,y=res))+
  geom_point(color="red")+
  geom_hline(yintercept=0,color="blue")
```

\clearpage 

### 27. Install and then load `stargazer` to output your regression into a table. For simplicity, set `type=text` for now. Verify where everything is that you found for question \#24. 

```{r}
#install.packages("stargazer") # install for first time
library("stargazer")
stargazer(reg,type="text")
```

