---
title: "R Practice 4 Answers"
author: "Ryan Safner"
date: "11/15/2018"
output: pdf_document
---

# Intro to `dplyr` syntax

### 1. Load the `gapminder` and `tidyverse` packages. 

```{r}
suppressPackageStartupMessages(library("tidyverse"))
library("gapminder")
```

## `select()`

### 1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

```{r}
select(gapminder, c(year, lifeExp, country))

# using the pipe

gapminder %>%
  select(c(year, lifeExp, country))
```

### 2. Select all variables, from `country` to `lifeExp`. 

```{r}

```

### 3. Select all variables, except `pop`.

```{r}
select(gapminder, -pop)

# using the pipe

gapminder %>%
  select(-pop)
```

### 4. Rename `continent` to `cont`.

```{r}
rename(gapminder, cont=continent)

# using the pipe

gapminder %>%
  rename(cont=continent)
```

## `arrange()`

### 1. Order by year.

```{r}
gapminder %>%
  arrange(year)
```

### 2. Order by year, in descending order.

```{r}
gapminder %>%
  arrange(desc(year))
```

### 3. Order by year, then by life expectancy.

```{r}
gapminder %>%
  arrange(year, lifeExp)
```

## Piping, `%>%`

Note: think of `%>%` as the word "then"!

### 1. Subset your data to look only at `year`, `gdpPercap`, and `country` in the year 1997, for countries that have a `gdpPercap` greater than 20,000, and order them alphabetically. 

```{r}
gapminder %>%
  select(year, gdpPercap, country) %>%
  filter(year==1997,
         gdpPercap>20000) %>%
  arrange(country)
```

Combine `select()` Task 1 with `arrange()` Task 3.

## `filter()`

### 1. Only take data with population greater than 1 billion.

```{r}
gapminder %>%
  filter(pop>1000000000)
```

### 2. Of those, only look at data from China. 

```{r}
gapminder %>%
  filter(pop>1000000000) %>%
  filter(country=="India")
```

## `mutate()`

### 1. Make a new variable that is `GDP` instead of `gdpPercap` (multiply `gdpPercap` by `pop`).

```{r}
gapminder %>%
  mutate(GDP=gdpPercap*pop)
```

### 2. Make a new variable for `gdpPercap` that is in millions.  

```{r}
gapminder %>%
  mutate(GDP=gdpPercap*pop) %>%
  mutate(GDPm=(GDP/1000000))
```

### 3. Make a new population variable that is the population in millions. 

```{r}
gapminder %>%
  mutate(popm=pop/1000000)
```

## `summarize()`

### 1. Get the average GDP per capita

```{r}
gapminder %>%
  summarize(mean(gdpPercap))
```

### 2. Get the number of observations, average, minimum, maximum, and standard deviation for GDP per capita.

```{r}
gapminder %>%
  summarize(Obs=n(),
            Average=mean(gdpPercap),
            Minimum=min(gdpPercap),
            Maximum=max(gdpPercap),
            SD=sd(gdpPercap))
```

### 3. Get the average for GDP per capita, Life expectancy, and population

```{r}
gapminder %>%
  summarize(Average_GDPcapita=mean(gdpPercap),
            Average_LE=mean(lifeExp),
            Average_pop=mean(pop))
```

## `group_by()`

### 1. Track the change in average GDP per capita over time. Hint, first group by year. 

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(Average_GDPcapita=mean(gdpPercap))
```

### 2. Get the average GDP per capita by continent.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(Average_GDPcapita=mean(gdpPercap))
```

### 3. You can group by multiple groups. Try getting the average GDP per capita by year by continent. Hint: do `year` first, if you do `continent` first, there are no years to group by! 

```{r}
gapminder %>%
  group_by(year, continent) %>%
  summarize(Average_GDPcapita=mean(gdpPercap))
```


