---
title: 'R Workshop #1'
author: "Ryan Safner"
date: "11/27/2018"
output: 
  beamer_presentation:
    #theme: "metropolis"
    incremental: true 
    fig_caption: yes
    toc: true 
    slide_level: 3
    includes:
      in_header: ../Lectures/header.tex
    keep_tex: no 
    latex_engine: xelatex #required for Fira Sans font to render properly 

classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library("tidyverse"))
```


# Subsetting Data 

### Subsetting Data: Matrix Analogy

- `data.frame` is a type of `matrix`: each cell is \alert{indexed} by its `[row #, column #]`

\scriptsize 

```{r}
m<-matrix(c("a","b","c","d","e","f"),nrow=2)
m
```

\columnsbegin

\column{.3\textwidth}

\scriptsize 

\onslide<2->

- Subset a **specific row**:
```{r}
m[2,]
```

\column{.3\textwidth}

\scriptsize 

\onslide<3->

- Subset a **specific column**:
```{r}
m[,3]
```

\column{.3\textwidth}

\scriptsize 

\onslide<4->

- Subset a **specific element**:
```{r}
m[2,3]
```

\columnsend

### Subsetting Data

\begin{figure}
\includegraphics{subsetting}
\end{figure}


### Subsetting Data: Data Frames

- We can do the same thing for `data.frame`s: 

```{r}
df<-data.frame(Nums=c(1,2,3,4,5),
               Lets=c("a","b","c","d","e"))
df
```

```{r, echo=F}
df$Lets<-as.character(df$Lets)
```

### Subsetting Data: Data Frames

- Can also subset a `data.frame` by position:

\scriptsize 

\center

```{r}
df
```

\columnsbegin

\column{.3\textwidth}

\scriptsize 

\onslide<2->

- Subset a **specific row (observation)**:
```{r}
df[2,]
```

\column{.3\textwidth}

\scriptsize 

\onslide<3->

- Subset a **specific column (variable)**:
```{r}
df[,2]
```

\column{.3\textwidth}

\scriptsize 

\onslide<4->

- Subset a **specific value**:
```{r}
df[2,2]
```

\columnsend

### Subsetting Data: Data Frames II 

- The nice thing about data frames is that instead of remembering the order of columns, we have the `names` of columns

```{r}
df
names(df)
```

- We subset by variable name with the `$` sign instead of brackets! 

```{r}
df$Nums

df$Lets
```

### Subsetting Conditionally

- We often want to subset a `data.frame` based on a \alert{condition}
    - e.g. look only at **observations for which `Nums` are larger than 2**
- Can use original brackets `[]` to pick by **rows** (observations) for which `Num>2`

\columnsbegin

\column{.3\textwidth}

\scriptsize 

\onslide<4->

- If we want **all** columns (variables)
```{r}
df[df$Nums>2,]
```

\column{.3\textwidth}

\scriptsize 

\onslide<5->

- If we only want column 1 (`"Nums"`)
```{r}
df[df$Nums>2,1]
```

\column{.3\textwidth}

\scriptsize 

\onslide<6->

- If we only want column 2 (`"Lets"`)
```{r}
df[df$Nums>2,2]
```

\columnsend


### Subsetting Conditionally: `subset()`

- One faster way that gets us away from `[]` is `subset(df, condition)`
    - Keeps only values of `df` for which condition is `TRUE`

\onslide<3->
```{r}
subset(df, Nums>2)
```

### Subsetting Conditionally: `filter()` with `dplyr`

- `dplyr` makes this easier with `filter()`

```{r}
df %>%
  filter(Nums>2)
```

### Useful Conditionals 

| Condition | Description | Example(s) |
|-----------------------------|------------------------------|----------------------------|
| `>`       | Values greater than | `Num>2` | 
| `>=`      | Values greater than or equal to | `Num>=2` | 
| `==`      | Values equal to (put value in quotes if a character) | `Num==2`; `Let=="a"` | 
| `!=`      | Values are NOT equal to | `Num!=2`; `Let!="a"` |
| `cond.1 & cond.2` | "AND": BOTH conditions must be met  | `Num>2 & Num<5` | 
| `cond.1 | cond.2` | "OR": Either one condition must be met | `Num>2 | Num<5` | 
| `%in% c()`    | Values are in a set of values defined in `c()` | `Num %in% c(1,2,3)` | 
| `!%in% c()`   | Values are NOT in defined set | `Num !%in% c(1,2,3)` | 

# Dealing with Missing Data

### Missing Data: `NA`

- If any observation is missing a value of a variable, it will show up as `NA`

```{r}
x<-c(1,2,NA,4,5)
y<-c("a",NA,"c","d","e")
df<-data.frame(x,y)

df

```

### `NA`s Propagate...

- Missing data propagates and will ruin many functions you run on it 

```{r}
mean(df$x)
sd(df$x)
sum(df$x)
```

### `NA` Strategy

- Several strategies to combat `NA`s
1. If looking at one variable:
    - Keep only observations for which there are no `NA`s 
```{r}
# with base R

df1<-df[!is.na(df$x),] # drop all observations for which there is NA for x 
df1

# alternatively, with dplyr

df %>%
  filter(!is.na(x)) # keep only observations that are NOT (the !) NA's 
```

- Notice this keeps the `df$y` with a `NA`

### `NA` Strategy II

2. Drop *all* observations that have some missing value across *any* variable with `na.omit(df)`
    - Often too extreme, may end up throwing out a lot of useful data! 
    
```{r}
df2<-na.omit(df) # drop any row that has any NA value for any variable 
df2
```

### `NA` Strategy III

3. Most functions have a `NA` option built in
    - Add "`, na.rm=TRUE`" inside any function's `()` to simply *ignore* all observations with `NA`s

```{r}
mean(df$x, na.rm=TRUE)
sd(df$x, na.rm=TRUE)
sum(df$x, na.rm=TRUE)
```


# Managing Your Workflow with `R Projects`

### Using `RProj` Projects

\begin{figure}
\includegraphics[height=2in]{workflowgithub.png}
\caption*{Go to \href{http://github.com/ryansafner/workflow}{github.com/ryansafner/workflow} and follow the instructions!}
\end{figure}

